---
title: "PD, microbiome and host genetics"
author: "Giacomo Antonello, Zachary D. Wallen, Chad Murchinson, Levi Waldron, Haydeh Payami"
date: "`r Sys.date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: hide
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
```

```{r}
library(tidyverse)
library(readxl)
# ideally this should one day be a standalone package
source("~/bin/biobakeryUtils/wrangle_metaphlan.R")
source("~/bin/biobakeryUtils/complete_unknown_taxonomy.R")
source("~/bin/biobakeryUtils/wrangle_humann.R")
if(!("gwasrapidd" %in% installed.packages())) install.packages("gwasrapidd")
if(!("biomaRt" %in% installed.packages())) BiocManager::install("biomaRt")

library(DT)
library(kableExtra)
library(ggpubr)

library(MultiAssayExperiment)
library(MaAslin2)
library(ANCOMBC)
```

# Directory and data preparation

First thing is clone the repository I made for the Payami Lab and add make sure 
that the `.gitignore` file has exceptions for unwanted data sharing into the 
repository. `NB: This has to be done manually`

```{bash}
cd ~

git clone git@github.com:ASAP-MAC/PayamiH.git

cd PayamiH
```

## Work directories and genetic data

1 - Data directories generation

```{bash setup data directory, eval=FALSE}
# TOPMed data
mkdir -p Data/TOPMed_imputed_genotypes
cp -r /data/project/payami-lab/payamilab/TOPMed_imputed_genotypes_2022/* Data/TOPMed_imputed_genotypes

# Microbiome data
mkdir -p Data/Zenodo
wget https://zenodo.org/records/7246185/files/Source_Data_24Oct2022.xlsx?download=1 -O /Data/Zenodo/Source_Data_24Oct2022.xlsx

```

2 - bcftools setup

```{bash setup micromamba, eval=FALSE}
cd ~ 

curl -Ls https://micro.mamba.pm/api/micromamba/linux-64/latest | tar -xvj micromamba/bin
./bin/micromamba shell init -s bash -r ~/micromamba

micromamba config append channels bioconda

micromamba create -n geneticTools -c bioconda

micromamba activate geneticTools

micromamba install bcftools

```

3 - Indexing of VCF file for fast querying (necessary for bcftools)

```{bash tabix vcf files, eval=FALSE}
cd ~/PayamiH/microbiome_genetics_PD/Data/TOPMed_imputed_genotypes/

for chr in `ls *.dose.vcf.gz`; do sbatch -c1 --mem 8G -pexpress --wrap "bcftools index ${chr}"; done
```

4 - Extract samples list to query with `bcftools query`, this is done considering the following:

  1. The only samples needed for the analysis are those beginning with `D`
  2. Keep only samples from point 1 that present across all VCF files. This is 
  redundant but done anyway.

```{bash, eval=FALSE}
mkdir -p ~/PayamiH/microbiome_genetics_PD/Data/sampleNames/

cd ~/PayamiH/microbiome_genetics_PD/Data/

find TOPMed_imputed_genotypes_2022/ -type f -name "*.vcf.gz" -exec bcftools query -l {} \; >> sampleNames/sampleNamesAll.txt

cat sampleNames/sampleNamesAll.txt | egrep "^D" | sort | uniq -c | sed -e "s/^[\ ]*//" | tr " " "\t" | awk '{if ($1 == 23) print $2}' > sampleNames/sampleNamesPD.txt

cat sampleNames/sampleNamesPD.txt | tr "\n" "," | sed "s/\,$//g" > sampleNames/sampleNamesPD_oneLine.txt
```

The last step is important for bcftools to work properly in querying samples. 

# Project setup, PD genetics and gut microbiome

```{bash}
cd ~/PayamiH

mkdir project_1_MicroGenPD/output/

```


Now we should download the genetic variants that we are interested in, being
those associated with `PD`.

```{r, eval=interactive()}
# retrieve associations with GWAS catalog traits
parkinson_gwasCatalog.gwasrapidd <- gwasrapidd::get_associations(efo_id = "MONDO_0005180")

parkinson_variants_gwas_catalog.df <- full_join(
  select(parkinson_gwasCatalog.gwasrapidd@associations, 
         association_id, or_per_copy_number, beta_unit),
  select(parkinson_gwasCatalog.gwasrapidd@risk_alleles, 
         association_id, variant_id, risk_allele), 
  by = "association_id") |> 
  full_join(parkinson_gwasCatalog.gwasrapidd@genes |>  
              group_by(association_id) |> 
              summarise(
                genes = paste(gene_name, collapse = ";")
              ), 
            by = "association_id")

ensembl <- biomaRt::useEnsembl(biomart = "snps", dataset = "hsapiens_snp")

rsid_to_CHR_POS <- biomaRt::getBM(
  attributes = c("refsnp_id", "chr_name", "chrom_start", "chrom_end", "allele"), 
      filters = "snp_filter", 
      values = parkinson_variants_gwas_catalog.df$variant_id, 
      mart = ensembl) |> 
  rename(
    variant_id = refsnp_id
  ) |>
  #keep only well defined chromosome scaffolds, including sex chromosome
  filter(
    !startsWith(chr_name, "H")
  ) |> 
  left_join(distinct(select(parkinson_variants_gwas_catalog.df,variant_id, genes)), by = "variant_id") |> 
  mutate(
    chr_name = factor(paste0("chr", chr_name), levels = as.character(paste0("chr", c(1:22, "X", "Y"))))
  ) |> 
  select(CHROM = chr_name, CHROM_START = chrom_start, CHROM_END = chrom_end, ALLELE = allele, GENE = genes, RSID = variant_id) 

write_tsv(rsid_to_CHR_POS, file = "../Data/GWAS_catalog_PD_variants/PD_variants_GWAS_catalog.tsv")

lapply(unique(rsid_to_CHR_POS$CHROM), function(chr) 
  writeLines(text = paste0(filter(rsid_to_CHR_POS, CHROM == chr)$CHROM, ":", filter(rsid_to_CHR_POS, CHROM == chr)$CHROM_START,"-", filter(rsid_to_CHR_POS, CHROM == chr)$CHROM_END), con = paste0("../Data/GWAS_catalog_PD_variants/", chr, "_variants_Tabix.txt")))

datatable(rsid_to_CHR_POS)
```

```{bash, eval=FALSE}
# from the PayamiH as base directory, do the following
cd Data/GWAS_catalog_PD_variants 

for chr in `ls | grep -oE "^(chr[0-9]*|chrX)"`; do bcftools query -s `cat ../sampleNames/sampleNamesPD_oneLine.txt` -H -f '%CHROM\t%POS\t%REF\t%ALT\t%INFO/R2[\t%GT]\n' -r `cat ${chr}_variants_Tabix.txt|tr "\n" "," | sed "s/\,$//g"`  ../TOPMed_imputed_genotypes/${chr}.dose.vcf.gz | sed -e "s/\#//g"| sed "s/\:GT//g" | sed -e "s/\[[0-9]*\]//g" > queried_variants_PD/${chr}_pd_variants.txt; done
```

## Retrieve PD associated variants in SNCA and HLA genes

Additionally, they must be available in WallenZD

```{r}
library(data.table)

gene_names <- fread("Data/GWAS_catalog_PD_variants/PD_variants_GWAS_catalog.tsv") %>% 
  select(CHROM, POS = CHROM_START, GENE, RSID) %>% 
  # force unique chrom and pos and group Genes and rsid together
  group_by(CHROM, POS) %>% 
  reframe(
    GENE = paste(GENE, collapse = ";"),
    RSID = paste(RSID, collapse = ";")
  )

queried_files.list <- list.files(path = "Data/GWAS_catalog_PD_variants/queried_variants_PD/", full.names = T) %>% 
  sapply(fread, USE.NAMES = TRUE, simplify = FALSE) %>% 
  lapply(left_join, gene_names, by = c("CHROM", "POS")) %>% 
  lapply(relocate, RSID, GENE, .after = ALT) %>% 
  lapply(function(x) set_names(x, str_remove(colnames(x), "_.*")))

PD_variants_genotypes_carriers <- reduce(queried_files.list, rbind.data.frame)

SNCA_HLA_variants_only <- filter(PD_variants_genotypes_carriers, grepl("SNCA|HLA", GENE)) %>% 
  as.data.frame()
rownames(SNCA_HLA_variants_only) <- paste(SNCA_HLA_variants_only$CHROM, SNCA_HLA_variants_only$POS, sep = "_")

SNCA_HLA_variants_only <- select(SNCA_HLA_variants_only, starts_with("D")) %>% 
  t()
```

At this step, we have generated a set of genotypes to test against the microbiome
as found in the GWAS catalog and our genotypes. It is a `samples x genotypes` matrix.

## Microbiome data preparation into `MultiAssayExperiment`

```{r}

theme_set(theme_light())
# load and wrangle available data

# 1 - metaphlan relative abundances and estimated counts
WallenZD_2022_metaphlan_relab <- read_xlsx("Data/Zenodo/Source_Data_24Oct2022.xlsx", sheet = "metaphlan_rel_ab")
WallenZD_2022_metaphlan_counts <- read_xlsx("Data/Zenodo/Source_Data_24Oct2022.xlsx", sheet = "metaphlan_counts")
# 1.1 - clean metaphlan tables
WallenZD_2022_metaphlan_relab_wrangled <- wrangle_metaphlan(WallenZD_2022_metaphlan_relab, "Species")
WallenZD_2022_metaphlan_counts_wrangled <- wrangle_metaphlan(WallenZD_2022_metaphlan_counts, "Species")
WallenZD_2022_metaphlan_counts_wrangled$profiles <- round(WallenZD_2022_metaphlan_counts_wrangled$profiles, 0)

# 2 - HUMaNN profiles
# 2.1 KO tables
WallenZD_2022_humann_KO <- read_xlsx("Data/Zenodo/Source_Data_24Oct2022.xlsx", sheet = "humann_KO_group_counts") |> 
  wrangle_humann(input_type = "ko")
WallenZD_2022_humann_KO$profiles <- round(WallenZD_2022_humann_KO$profiles, 0)
WallenZD_2022_humann_pathways <- read_xlsx("Data/Zenodo/Source_Data_24Oct2022.xlsx", sheet = "humann_pathway_counts") |> wrangle_humann(input_type = "pathways")

# 2.2 - Pathways
WallenZD_2022_humann_pathways$profiles <- round(WallenZD_2022_humann_pathways$profiles, 0)

# 3 - sample-participant metadata
WallenZD_2022_metadata <- read_xlsx("Data/Zenodo/Source_Data_24Oct2022.xlsx", sheet = "subject_metadata") |> 
  as.data.frame()
rownames(WallenZD_2022_metadata) <- WallenZD_2022_metadata$sample_name

# reorder colnames so they all match the metadata
samples_intersect <- purrr::reduce(list(WallenZD_2022_metaphlan_relab_wrangled$profiles,
WallenZD_2022_metaphlan_counts_wrangled$profiles,
WallenZD_2022_humann_KO$profiles,
WallenZD_2022_humann_pathways$profiles) |> lapply(colnames), intersect) 

WallenZD_2022_metaphlan_relab_wrangled$profiles <- WallenZD_2022_metaphlan_relab_wrangled$profiles[, samples_intersect]
WallenZD_2022_metaphlan_counts_wrangled$profiles <- WallenZD_2022_metaphlan_counts_wrangled$profiles[, samples_intersect]
WallenZD_2022_humann_KO$profiles <- WallenZD_2022_humann_KO$profiles[, samples_intersect]
WallenZD_2022_humann_pathways$profiles <- WallenZD_2022_humann_pathways$profiles[, samples_intersect]

sample_metadata.df <- WallenZD_2022_metadata[samples_intersect,]

```

```{r}
library(MultiAssayExperiment)

wallenZD_summarizedExperiments.list <- list(
  metaphlan_RelAbund.sa = SummarizedExperiment(
    assays = list("metaphlan_RelAbund" = WallenZD_2022_metaphlan_relab_wrangled$profiles),
    rowData =  WallenZD_2022_metaphlan_relab_wrangled$taxonomies
  ),
  
  metaphlan_Counts.sa = SummarizedExperiment(
    assays = list("metaphlan_Counts" = WallenZD_2022_metaphlan_counts_wrangled$profiles),
    rowData =  WallenZD_2022_metaphlan_counts_wrangled$taxonomies
  ),
  
  humann_ko.sa = SummarizedExperiment(
    assays = list("humann_ko" = WallenZD_2022_humann_KO$profiles),
    rowData =  WallenZD_2022_humann_KO$description
  ),
  
  humann_pathways.sa = SummarizedExperiment(
    assays = list("humann_pathways" = WallenZD_2022_humann_pathways$profiles),
    rowData =  WallenZD_2022_humann_pathways$description
  )
)

assays_sample_map <- data.frame(
  assay = factor(lapply(wallenZD_summarizedExperiments.list, assays) %>% 
    names() %>% 
    lapply(rep, length(samples_intersect)) %>% 
    reduce(c), levels = sapply(wallenZD_summarizedExperiments.list, assays) %>% 
    names()),
  primary = lapply(wallenZD_summarizedExperiments.list, colnames) %>% reduce(c),
  colname = lapply(wallenZD_summarizedExperiments.list, colnames) %>% reduce(c)
)


wallenZD_data.mae <- MultiAssayExperiment(
  experiments = wallenZD_summarizedExperiments.list,
  colData = sample_metadata.df,
  sampleMap = assays_sample_map
)

rm(list = grep("wallenZD_data\\.mae|PD_variants_genotypes_carriers|SNCA_HLA_variants_only", ls(), value = TRUE, invert = TRUE))
```

# Differentially abundant microbiome features with PD

```{r}
out.dir <- "analyses_and_results"
```

First differential abundance done on Maaslin2 

```{r, message=FALSE, warning=FALSE}
library(Maaslin2)
myMaaslin2 <- Maaslin2(
  input_data = microbiome_wide_datasets.list$metaphlan_relAb$profiles / 100,
  input_metadata = sample_metadata.df |> mutate(seqs_scaled = scale(total_sequences)),
  output = paste(out.dir, "/Maaslin2_div100"),
  fixed_effects = c('Case_status', 'collection_method', 'seqs_scaled'),
  min_prevalence = 0.05,
  normalization = 'NONE',
  max_significance = 0.05,
  standardize = FALSE,
  plot_heatmap = FALSE,
  plot_scatter = FALSE,
  heatmap_first_n = 0
)

maaslin2_results <- myMaaslin2$results |> 
  group_by(metadata) |> 
  mutate(FDR_grouped = p.adjust(pval, method = "BH")) |> 
  ungroup() |> 
  arrange(feature)|> 
  mutate(FC = 2^coef) |> 
  relocate(FC, .after = coef)

```

Second done on ANCOM-BC (not ANCOM-BC2)

```{r}
library(phyloseq, quietly = T, verbose = FALSE)

ps.counts <- phyloseq(
  otu_table(microbiome_wide_datasets.list$metaphlan_counts$profiles, taxa_are_rows = TRUE), 
  tax_table(as.matrix(microbiome_wide_datasets.list$metaphlan_counts$taxonomies)),
  sample_data(sample_metadata.df |> mutate(seqs_scaled = scale(total_sequences)))
  )

ancombc_ga <- ANCOMBC::ancombc(
  data = ps.counts,
  prv_cut = 0.05,
  formula="Case_status + collection_method + seqs_scaled",
  p_adj_method = "BH",
  verbose = FALSE, 
  alpha = 0.1
  )

dir.create("ANCOM-BC/", showWarnings = FALSE, recursive = TRUE)
saveRDS(ancombc_ga, file = "ANCOM-BC/ancombc_ga_replication.Rds")

ancombc_ga.df <- ancombc_ga$res |> 
  sapply(select, contains("Case_status")) |> 
  bind_cols() |> 
  rename_all(.funs = function(x) str_remove(x, "\\.Case_statusPD")) |> 
  mutate(
    taxon = ancombc_ga$res$lfc$taxon,
    FC = exp(lfc)
    ) |> 
  relocate(taxon, lfc, FC) |> 
  arrange(taxon)
```

Third, summarize results into one data frame with the same rationale used
in the Nat. Comms. paper (2022)

```{r}
# ANCOM-BC model was done and saved in Supplementary.Rmd 
tmp <- readRDS(file = "ANCOM-BC/ancombc_ga_replication.Rds")
ancombc_ga.df <- tmp$res |> 
  sapply(select, contains("Case_status")) |> 
  bind_cols() |> 
  rename_all(.funs = function(x) str_remove(x, "\\.Case_statusPD")) |> 
  mutate(
    taxon = tmp$res$lfc$taxon,
    FC = exp(lfc)
    ) |> 
  relocate(taxon, lfc, FC) |> 
  arrange(taxon)

rm(tmp)

wallen_full_results <- merge(
  maaslin2_results |> filter(metadata == "Case_status") |> select(Feature = feature, FDR = FDR_grouped, FC),
  ancombc_ga.df |> select(Feature = taxon, FDR = q_val, FC),
                   by='Feature', suffix=c('_maaslin','_ancombc')) |> 
  mutate(
    PD_association = ifelse(
      (round(FDR_maaslin, 1) <= 0.1 & FDR_ancombc < 0.05) | (round(FDR_ancombc, 1) <= 0.1 & FDR_maaslin < 0.05),
      "associated",
      "not associated"
    ),
    PD_association_direction = case_when(
      PD_association == "associated" & FC_maaslin > 1 & FC_ancombc > 1 ~ "enriched",
      PD_association == "associated" & FC_maaslin < 1 & FC_ancombc < 1 ~ "depleted",
      PD_association == "not associated" ~ "not associated",
      TRUE ~ "opposite directions"
    )
  )

datatable(wallen_full_results)
```


# OLD CODE (To be recicled up) - PD-associated genetic variants

Retrieval was done using `gwasrapidd` and `biomaRt` with the following method:

  1. Find associations reported for `Parkinson disease` as generic term, with 
  ontology name `MONDO_0005180`. This by default returns `rsID`s
  2. Retrieve chromosome and position with `biomaRt`
  
Next, these chromosomes and positions can be searched on the TOPMed imputed 
genotypes and carriers will be extracted with tabix. 

**Ideally, `bcftools` (`samtools`) should be used, but I can't find them installed
nor I can install neither with bash nor with python**, so now I working around 
it.



# Association between top PD associated taxa and genotypes

First challenge is defining top taxa. Haydeh chose based on signifincance 
theshoulds and correlation network of the taxa. The strategy is below




# R session

```{r}
sessionInfo()
```

