---
title: "PD, microbiome and host genetics"
author: "Giacomo Antonello, Zachary D. Wallen, Chad Murchinson, Levi Waldron, Haydeh Payami"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: hide
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
```

```{r}
# Check and install necessary CRAN packages
CRAN_required_packages <- c("tidyverse", "gwasrapidd", "DT", "kableExtra", "data.table", "ggpubr", "BiocManager", "readxl")

install.packages(
  CRAN_required_packages[!(CRAN_required_packages %in% installed.packages())] 
)

# Check and install necessary Bioconductor packages
BioC_required_packages <- c("Maaslin2", "ANCOMBC", "biomaRt", "mia", "phyloseq", "microbiome","SIAMCAT")

BiocManager::install(
  BioC_required_packages[!(BioC_required_packages %in% installed.packages())], ask = FALSE, update = TRUE 
)

# load basic data manipulation packagesß
library(tidyverse)
library(DT)
# ideally this should one day be a standalone package
source("~/bin/biobakeryUtils/wrangle_metaphlan.R")
source("~/bin/biobakeryUtils/complete_unknown_taxonomy.R")
source("~/bin/biobakeryUtils/wrangle_humann.R")

```

# Directory and data preparation

First thing is clone the repository I made for the Payami Lab and add make sure 
that the `.gitignore` file has exceptions for unwanted data sharing into the 
repository. `NB: This has to be done manually`

```{bash, eval=FALSE}
cd ~

git clone git@github.com:ASAP-MAC/PayamiH.git

cd PayamiH
```

## Work directories and genetic data

1 - Data directories generation

```{bash setup data directory, eval=FALSE}
# TOPMed data
mkdir -p Data/TOPMed_imputed_genotypes
cp -r /data/project/payami-lab/payamilab/TOPMed_imputed_genotypes_2022/* Data/TOPMed_imputed_genotypes

# Microbiome data
mkdir -p Data/Zenodo
wget https://zenodo.org/records/7246185/files/Source_Data_24Oct2022.xlsx?download=1 -O /Data/Zenodo/Source_Data_24Oct2022.xlsx

```

2 - bcftools setup

```{bash setup micromamba, eval=FALSE}
cd ~ 

curl -Ls https://micro.mamba.pm/api/micromamba/linux-64/latest | tar -xvj micromamba/bin
./bin/micromamba shell init -s bash -r ~/micromamba

micromamba config append channels bioconda

micromamba create -n geneticTools -c bioconda

micromamba activate geneticTools

micromamba install bcftools

```

3 - Indexing of VCF file for fast querying (necessary for bcftools)

```{bash tabix vcf files, eval=FALSE}
cd ~/PayamiH/Data/TOPMed_imputed_genotypes/

for chr in `ls *.dose.vcf.gz`; do sbatch -c1 --mem 8G -pexpress --wrap "bcftools index -c ${chr}"; done
```

4 - Extract samples list to query with `bcftools query`, this is done considering the following:

  1. The only samples needed for the analysis are those beginning with `D`
  2. Keep only samples from point 1 that present across all VCF files. This is 
  redundant but done anyway.

```{bash, eval=FALSE}
mkdir -p ~/PayamiH/Data/sampleNames/

cd ~/PayamiH/Data/

find TOPMed_imputed_genotypes_2022/ -type f -name "*.vcf.gz" -exec bcftools query -l {} \; >> sampleNames/sampleNamesAll.txt

cat sampleNames/sampleNamesAll.txt | egrep "^D" | sort | uniq -c | sed -e "s/^[\ ]*//" | tr " " "\t" | awk '{if ($1 == 23) print $2}' > sampleNames/sampleNamesPD.txt

cat sampleNames/sampleNamesPD.txt | tr "\n" "," | sed "s/\,$//g" > sampleNames/sampleNamesPD_oneLine.txt
```

The last step is important for bcftools to work properly in querying samples. 

# Project setup, PD genetics and gut microbiome

```{bash, eval=FALSE}
cd ~/PayamiH

mkdir -p project_1_MicroGenPD/output/

```


## Download all unique GWAS catalog variants with PD (based on `rsID`)

Now we should download the genetic variants that we are interested in, being
those associated with `PD`. Notice that there are duplicated `rsID`s depending
on how many times they have been entried, hence reported. We are interested in 
having them only once.

```{r, eval=FALSE}
# retrieve associations with GWAS catalog traits
parkinson_gwasCatalog.gwasrapidd <- gwasrapidd::get_associations(efo_id = "MONDO_0005180")

parkinson_variants_gwas_catalog.df <- full_join(
  select(parkinson_gwasCatalog.gwasrapidd@associations, 
         association_id, or_per_copy_number, beta_unit),
  select(parkinson_gwasCatalog.gwasrapidd@risk_alleles, 
         association_id, variant_id, risk_allele), 
  by = "association_id") %>% 
  # remove duplicated lines, preferring those where a rik allele has been reported
  arrange(risk_allele) %>%       
  distinct(variant_id, .keep_all = TRUE)

ensembl <- biomaRt::useEnsembl(biomart = "snps", dataset = "hsapiens_snp")

rsid_to_CHR_POS <- biomaRt::getBM(
  attributes = c("refsnp_id", "chr_name", "chrom_start", "chrom_end", "allele", "associated_gene"), 
      filters = "snp_filter", 
      values = parkinson_variants_gwas_catalog.df$variant_id, 
      mart = ensembl) %>% 
  rename(
    variant_id = refsnp_id
  ) %>%
  #keep only well defined chromosome scaffolds, including sex chromosome
  filter(
    !startsWith(chr_name, "H")
  ) %>% 
  mutate(
    chr_name = factor(paste0("chr", chr_name), levels = as.character(paste0("chr", c(1:22, "X", "Y"))))
  ) %>% 
  select(
    CHROM = chr_name, 
    CHROM_START = chrom_start, 
    CHROM_END = chrom_end, 
    ALLELE = allele, 
    RSID = variant_id,
    GENE = associated_gene
    ) %>% 
  # same trick, to be sure we have only variants with gene names associated, wherever possible
  arrange(rowSums(is.na(.))) %>%       
  distinct(RSID, .keep_all = TRUE)

write_tsv(rsid_to_CHR_POS, file = "../Data/GWAS_catalog_PD_variants/PD_variants_GWAS_catalog.tsv")

lapply(unique(rsid_to_CHR_POS$CHROM), function(chr) 
  writeLines(text = paste0(filter(rsid_to_CHR_POS, CHROM == chr)$CHROM, ":", filter(rsid_to_CHR_POS, CHROM == chr)$CHROM_START,"-", filter(rsid_to_CHR_POS, CHROM == chr)$CHROM_END), con = paste0("../Data/GWAS_catalog_PD_variants/", chr, "_variants_Tabix.txt")))

datatable(rsid_to_CHR_POS, caption = "Variants available in the GWAS catalog made unique based on the rsID. Additionally, Gene annotations were given preference over no annotation. Hence, they should be treated with caution. All annotations were done using the ensembl annotation resources in `biomaRt`.")
```

```{bash, eval=FALSE}
# from the PayamiH as base directory, do the following
cd Data/GWAS_catalog_PD_variants 

for chr in `ls | grep -oE "^(chr[0-9]*|chrX)"`; do bcftools query -s `cat ../sampleNames/sampleNamesPD_oneLine.txt` -H -f '%CHROM\t%POS\t%REF\t%ALT\t%INFO/R2[\t%GT]\n' -r `cat ${chr}_variants_Tabix.txt|tr "\n" "," | sed "s/\,$//g"`  ../TOPMed_imputed_genotypes/${chr}.dose.vcf.gz | sed -e "s/\#//g"| sed "s/\:GT//g" | sed -e "s/\[[0-9]*\]//g" > queried_variants_PD/${chr}_pd_variants.txt; done
```

## Retrieve PD associated variants in SNCA and HLA genes in Wallen genotypes

Additionally, they must be available in WallenZD

```{r}
gwasCatalog_variants_All <- data.table::fread("../Data/GWAS_catalog_PD_variants/PD_variants_GWAS_catalog.tsv", data.table = FALSE) %>% 
  select(CHROM, POS = CHROM_START, GENE, RSID) %>% 
  # force unique chrom and pos and group Genes and rsid together, this is redundant
  group_by(CHROM, POS) %>% 
  reframe(
    GENE = paste(GENE, collapse = ";"),
    RSID = paste(RSID, collapse = ";")
  ) %>% as.data.frame()

gwasCatalog_variants_in_WallenZD.df <- list.files(path = "../Data/GWAS_catalog_PD_variants/queried_variants_PD/", full.names = T) %>% 
  sapply(data.table::fread, USE.NAMES = TRUE, simplify = FALSE) %>% 
  lapply(inner_join, gwasCatalog_variants_All, by = c("CHROM", "POS")) %>% 
  lapply(relocate, RSID, GENE, .after = ALT) %>% 
  lapply(function(x) set_names(x, str_remove(colnames(x), "_.*"))) %>% 
  purrr::reduce(rbind) %>% 
  as.data.frame()

# summary table of variants found in the dataset
nVars_by_genes_All <- gwasCatalog_variants_All %>% 
  mutate(
    `SNCA or HLA` = grepl("SNCA|HLA", GENE), 
    `LRRK2 or GBA` = grepl("LRRK2|GBA", GENE), 
    OTHER = !grepl("SNCA|HLA|GBA|LRRK2", GENE)) %>% 
  .[,(ncol(.)-2):ncol(.)] %>% 
  colSums()

nVars_by_genes_Wallen <- gwasCatalog_variants_in_WallenZD.df %>% 
  mutate(
    `SNCA or HLA` = grepl("SNCA|HLA", GENE), 
    `LRRK2 or GBA` = grepl("LRRK2|GBA", GENE), 
    OTHER = !grepl("SNCA|HLA|GBA|LRRK2", GENE)) %>% 
  .[,(ncol(.)-2):ncol(.)] %>% 
  colSums()

cbind(c("Tot" = sum(nVars_by_genes_All), nVars_by_genes_All), 
      c("Tot" = sum(nVars_by_genes_Wallen), nVars_by_genes_Wallen)) %>% 
  
  magrittr::set_colnames(c("GWAS Catalog", "WallenZD")) %>% 
  kableExtra::kbl(caption = "Comparison of genetic variants associated with PD found in the GWAS catalog and in Wallen imputed genotypes") %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::add_header_above(c(" " = 1, "N. unique variants associated with PD" = 2))

```


At this step, we have generated a set of genotypes to test against the microbiome
as found in the GWAS catalog and our genotypes. It is a `samples x genotypes` matrix.

## WallenZD data preparation into `MultiAssayExperiment`

```{r}

theme_set(theme_light())
# load and wrangle available data

# 1 - metaphlan relative abundances and estimated counts
WallenZD_2022_metaphlan_relab <- readxl::read_xlsx("../Data/Zenodo/Source_Data_24Oct2022.xlsx", sheet = "metaphlan_rel_ab")
WallenZD_2022_metaphlan_counts <- readxl::read_xlsx("../Data/Zenodo/Source_Data_24Oct2022.xlsx", sheet = "metaphlan_counts")
# 1.1 - clean metaphlan tables
WallenZD_2022_metaphlan_relab_wrangled <- wrangle_metaphlan(WallenZD_2022_metaphlan_relab, "Species")
WallenZD_2022_metaphlan_counts_wrangled <- wrangle_metaphlan(WallenZD_2022_metaphlan_counts, "Species")
WallenZD_2022_metaphlan_counts_wrangled$profiles <- round(WallenZD_2022_metaphlan_counts_wrangled$profiles, 0)

# 2 - HUMaNN profiles
# 2.1 KO tables
WallenZD_2022_humann_KO <- readxl::read_xlsx("../Data/Zenodo/Source_Data_24Oct2022.xlsx", sheet = "humann_KO_group_counts") %>% 
  wrangle_humann(input_type = "ko")
WallenZD_2022_humann_KO$profiles <- round(WallenZD_2022_humann_KO$profiles, 0)
WallenZD_2022_humann_pathways <- readxl::read_xlsx("../Data/Zenodo/Source_Data_24Oct2022.xlsx", sheet = "humann_pathway_counts") %>% wrangle_humann(input_type = "pathways")

# 2.2 - Pathways
WallenZD_2022_humann_pathways$profiles <- round(WallenZD_2022_humann_pathways$profiles, 0)

# 3 - sample-participant metadata
WallenZD_2022_metadata <- readxl::read_xlsx("../Data/Zenodo/Source_Data_24Oct2022.xlsx", sheet = "subject_metadata") %>% 
  as.data.frame()
rownames(WallenZD_2022_metadata) <- WallenZD_2022_metadata$sample_name

# reorder colnames so they all match the metadata
samples_intersect <- purrr::reduce(list(WallenZD_2022_metaphlan_relab_wrangled$profiles,
WallenZD_2022_metaphlan_counts_wrangled$profiles,
WallenZD_2022_humann_KO$profiles,
WallenZD_2022_humann_pathways$profiles) %>% lapply(colnames), intersect) 

WallenZD_2022_metaphlan_relab_wrangled$profiles <- WallenZD_2022_metaphlan_relab_wrangled$profiles[, samples_intersect]
WallenZD_2022_metaphlan_counts_wrangled$profiles <- WallenZD_2022_metaphlan_counts_wrangled$profiles[, samples_intersect]
WallenZD_2022_humann_KO$profiles <- WallenZD_2022_humann_KO$profiles[, samples_intersect]
WallenZD_2022_humann_pathways$profiles <- WallenZD_2022_humann_pathways$profiles[, samples_intersect]

sample_metadata.df <- WallenZD_2022_metadata[samples_intersect,] %>% 
  mutate(seqs_scaled = scale(total_sequences))

```

```{r}
# manipulate genetics data a bit, to make it more usable in the SummarizedExperiment
# context

tmp <- gwasCatalog_variants_in_WallenZD.df
samples_intersect <- intersect(colnames(tmp), rownames(sample_metadata.df))
metadata_cols_genetics <- grep("^D", colnames(tmp), value = TRUE, invert = T)
variants_made_unique <- make.unique(tmp$RSID)
rownames(tmp) <- variants_made_unique

genetics_matrix<- as.matrix(tmp[variants_made_unique, samples_intersect])
genetics_metadata <- tmp[variants_made_unique, metadata_cols_genetics]
sample_metadata <- sample_metadata.df[samples_intersect,]

library(MultiAssayExperiment, quietly = TRUE, verbose = FALSE)

wallenZD_summarizedExperiments.list <- list(
  genotypes.se = SummarizedExperiment(
    assays = list("genotypes.se" = genetics_matrix),
    rowData = genetics_metadata,
  colData = sample_metadata
  ),
  
  metaphlan_RelAbund.se = SummarizedExperiment(
    assays = list("metaphlan_RelAbund" = WallenZD_2022_metaphlan_relab_wrangled$profiles),
    rowData =  WallenZD_2022_metaphlan_relab_wrangled$taxonomies,
    colData = sample_metadata.df
  ),
  
  metaphlan_Counts.se = SummarizedExperiment(
    assays = list("metaphlan_Counts" = WallenZD_2022_metaphlan_counts_wrangled$profiles),
    rowData =  WallenZD_2022_metaphlan_counts_wrangled$taxonomies,
    colData = sample_metadata.df
  ),
  
  humann_ko.se = SummarizedExperiment(
    assays = list("humann_ko" = WallenZD_2022_humann_KO$profiles),
    rowData =  WallenZD_2022_humann_KO$description,
    colData = sample_metadata.df
  ),
  
  humann_pathways.se = SummarizedExperiment(
    assays = list("humann_pathways" = WallenZD_2022_humann_pathways$profiles),
    rowData =  WallenZD_2022_humann_pathways$description,
    colData = sample_metadata.df
  )
)

assays_vec <- lapply(wallenZD_summarizedExperiments.list, colnames) %>% 
  lapply(length)

assays_vec <- lapply(names(assays_vec), function(x) rep(x, assays_vec[[x]])) %>% 
  purrr::reduce(c) %>% 
  factor(levels = names(assays_vec))

assays_sample_map <- data.frame(
  assay = assays_vec,
  primary = lapply(wallenZD_summarizedExperiments.list, colnames) %>% purrr::reduce(c),
  colname = lapply(wallenZD_summarizedExperiments.list, colnames) %>% purrr::reduce(c)
)

wallenZD_data.mae <- MultiAssayExperiment(
  experiments = wallenZD_summarizedExperiments.list,
  sampleMap = assays_sample_map,
  colData = sample_metadata.df
)
```


```{r clean environment and memory for next analyses, echo=FALSE}
rm(list = grep("wallenZD_data\\.mae", ls(), value = TRUE, invert = TRUE))
gc()
save.image(file = "WallenFullDataReady.RData")

```

# Differentially abundant microbiome features with PD

First differential abundance done on Maaslin2

```{r, message=FALSE, warning=FALSE, include=FALSE}
dir.create("output/DiffAbundance/Maaslin2/", recursive = TRUE, showWarnings = FALSE)

myMaaslin2 <- Maaslin2::Maaslin2(
  input_data = assay(wallenZD_data.mae, "metaphlan_RelAbund.se") / 100,
    input_metadata = as.data.frame(colData(wallenZD_data.mae[["metaphlan_RelAbund.se"]])),
  output = "output/DiffAbundance/Maaslin2/",
  fixed_effects = c('Case_status', 'collection_method', 'seqs_scaled'),
  min_prevalence = 0.05,
  normalization = 'NONE',
  max_significance = 0.05,
  standardize = FALSE,
  plot_heatmap = FALSE,
  plot_scatter = FALSE,
  heatmap_first_n = 0
)

maaslin2_results.df <- myMaaslin2$results %>% 
  group_by(metadata) %>% 
  mutate(FDR_grouped = p.adjust(pval, method = "BH")) %>% 
  ungroup() %>% 
  arrange(feature)%>% 
  mutate(FC = 2^coef) %>% 
  relocate(FC, .after = coef)

rm(myMaaslin2)
```

Second done on ANCOM-BC (not ANCOM-BC2)

```{r}
library(phyloseq, quietly = T, verbose = FALSE)

ps.counts <- tryCatch(
  expr = mia::convertToPhyloseq(wallenZD_data.mae[["metaphlan_Counts.se"]], assay.type = "metaphlan_Counts"),
  error = function(e)
    return(
      phyloseq(
        otu_table(
          assay(wallenZD_data.mae, "metaphlan_Counts.se"),
          taxa_are_rows = TRUE
        ),
        tax_table(as.matrix(rowData(wallenZD_data.mae[["metaphlan_Counts.se"]]))),
        sample_data(as.data.frame(colData(wallenZD_data.mae[["metaphlan_Counts.se"]])))
      )
      )
  )

ancombc_results_raw <- ANCOMBC::ancombc(
  data = ps.counts,
  prv_cut = 0.05,
  formula = "Case_status + collection_method + seqs_scaled",
  p_adj_method = "BH",
  verbose = FALSE, 
  alpha = 0.1, 
  n_cl = 4
  )

dir.create("output/DiffAbund/ANCOMBC/", showWarnings = FALSE, recursive = TRUE)
saveRDS(ancombc_results_raw, file = "output/DiffAbund/ANCOMBC/ancombc_results_raw.Rds")

ancombc_results_clean.df <- ancombc_results_raw$res %>% 
  sapply(select, contains("Case_status")) %>% 
  bind_cols() %>% 
  rename_all(.funs = function(x) str_remove(x, "\\.Case_statusPD")) %>% 
  mutate(
    taxon = ancombc_results_raw$res$lfc$taxon,
    FC = exp(lfc)
    ) %>% 
  relocate(taxon, lfc, FC) %>% 
  arrange(taxon)

rm(ps.counts, ancombc_results_raw)
```

Third, summarize results into one data frame with the same rationale used
in the Nat. Comms. paper (2022)

```{r}
wallen_full_results.df <- merge(
  maaslin2_results.df %>% filter(metadata == "Case_status") %>% select(Feature = feature, FDR = FDR_grouped, FC),
  ancombc_results_clean.df %>% select(Feature = taxon, FDR = q_val, FC),
                   by='Feature', suffix=c('_maaslin','_ancombc')) %>% 
  mutate(
    PD_association = ifelse(
      (round(FDR_maaslin, 1) <= 0.1 & FDR_ancombc < 0.05) | (round(FDR_ancombc, 1) <= 0.1 & FDR_maaslin < 0.05),
      "associated",
      "not associated"
    ),
    PD_association_direction = case_when(
      PD_association == "associated" & FC_maaslin > 1 & FC_ancombc > 1 ~ "enriched",
      PD_association == "associated" & FC_maaslin < 1 & FC_ancombc < 1 ~ "depleted",
      PD_association == "not associated" ~ "not associated",
      TRUE ~ "opposite directions"
    )
  )

wallen_full_results.df %>% 
  filter(PD_association != "not associated") %>% 
  arrange(desc(abs(FC_maaslin + FC_ancombc))) %>% 
  mutate_if(is.numeric, .funs = function(col) ifelse(abs(col) < 0.001, format(col, digits = 3, scientific = TRUE), format(col, digits = 3))) %>% 
  datatable()
```

## Definition of microbiome subtypes based on Paper (SparCC network)

```{r, eval=FALSE}

control_samples <- colData(wallenZD_data.mae)$sample_name[colData(wallenZD_data.mae)$Case_status == "Control"]
PD_samples <- colData(wallenZD_data.mae)$sample_name[colData(wallenZD_data.mae)$Case_status == "PD"]

dir.create("output/SparCC/input_tables/bootstrapped_tables", recursive = TRUE, showWarnings = FALSE)
dir.create("output/SparCC/SparCC_results", recursive = TRUE, showWarnings = FALSE)

# write features table without all-zero columns formatted as SparCC likes them

assay(wallenZD_data.mae, "metaphlan_Counts.se") %>% 
  .[rowSums(.) > 0,PD_samples] %>% 
  rownames_to_column("OTU_id") %>% 
  filter(OTU_id != "UNKNOWN") %>% 
  write_tsv(file = "output/SparCC/input_tables/features_counts_PD.tsv")

assay(wallenZD_data.mae, "metaphlan_Counts.se") %>% 
  .[rowSums(.) > 0,control_samples] %>% 
  rownames_to_column("OTU_id") %>% 
  filter(OTU_id != "UNKNOWN") %>% 
  write_tsv(file = "output/SparCC/input_tables/features_counts_HC.tsv")


```

# Assocation between PD markers and PD genetics

```{r}
rm(list = ls())

load("WallenZD_data.mae")
library(tidyverse)
library(MultiAssayExperiment)
```

## Table of biomarkers

```{r}
biomarkers_medrxiv.df <- readxl::read_xlsx("")

datatable(biomarkers_medrxiv.df %>% select(-original_taxon_name))
```


Step 1, get list of genetic variants found in WallenZD under the `SNCA` and 
`HLA` gene annotations.

```{r}
# remove samples that were collected with swabs instead of the correct kit
samples_no_swab <- colData(wallenZD_data.mae)$sample_name[colData(wallenZD_data.mae)$collection_method != "swab"]
#
variants_SNCA_HLA <- rownames(rowData(wallenZD_data.mae[["genotypes.se"]])[grepl("HLA|SNCA", rowData(wallenZD_data.mae[["genotypes.se"]])$GENE),])

bugs_signatures <- biomarkers_medrxiv.df$original_taxon_name

genotypes_with_metadata <- assay(wallenZD_data.mae, "genotypes.se")[variants_SNCA_HLA,intersect(samples_no_swab, colnames(assay(wallenZD_data.mae, "genotypes.se")))] %>% 
  t() %>% 
  as.data.frame() %>%
  mutate_all(.funs = function(x) case_when(
    x == "0|0" ~ "HomRef",
    x == "1|1" ~ "HomAlt",
    TRUE ~ "Het"
  ) %>% factor(levels = c("HomRef", "Het", "HomAlt"))) %>% 
  rownames_to_column("sample_name") %>% 
  inner_join(as.data.frame(colData(wallenZD_data.mae)), by = "sample_name") %>% 
  filter(sample_name %in%  samples_no_swab) 

rownames(genotypes_with_metadata) <-  genotypes_with_metadata$sample_name


microbes_matrix <- as.matrix(assay(wallenZD_data.mae, "metaphlan_RelAbund.se"))[bugs_signatures,genotypes_with_metadata$sample_name] / 100

dir.create("output/genotypes_vs_bugs/SNCA_HLA", recursive = TRUE, showWarnings = FALSE)

maaslin_per_variant <- sapply(variants_SNCA_HLA,
       function(rsid)
         Maaslin2::Maaslin2(
           input_data = t(microbes_matrix),
           input_metadata = genotypes_with_metadata,
           output = "output/genotypes_vs_bugs/SNCA_HLA",
  fixed_effects = c(rsid, "Case_status", paste("Case_status", rsid, sep = "*"), "seqs_scaled", "Age_at_collection"),
  # min_prevalence = 0.05,
  normalization = 'NONE',
  max_significance = 0.05,
  standardize = FALSE,
  plot_heatmap = FALSE,
  plot_scatter = FALSE,
  heatmap_first_n = 0
  ), 
  simplify = FALSE, 
  USE.NAMES = TRUE
)

# group results based on cluster name
```


# alternative to clusters for biomarkers, variable importance in machine learning

```{r}
library(siamcat)
```

### R reimplementation of sparcc

SparCC is problematic on the Cheaha server, somehow the permutations step won't 
work.

```{r}
devtools::install_github("zdk123/SpiecEasi")

library(SpiecEasi)

PD_sparcc_base <- sparcc(
  assay(wallenZD_data.mae, "metaphlan_Counts.se") %>% 
  .[rowSums(.) > 0,PD_samples] %>% 
  rownames_to_column("OTU_id") %>% 
  filter(OTU_id != "UNKNOWN") %>% 
    column_to_rownames("OTU_id"), 
  iter = 100, th = 0.2)

PD_sparcc_boot <- sparccboot(data = assay(wallenZD_data.mae, "metaphlan_Counts.se") %>% 
  .[rowSums(.) > 0,PD_samples] %>% 
  rownames_to_column("OTU_id") %>% 
  filter(OTU_id != "UNKNOWN") %>% 
    column_to_rownames("OTU_id"), R = 1000)

saveRDS(PD_sparcc_boot, "PD_sparcc_boot.Rds")
#sparcc_pvalues <- pval.sparccboot(PD_sparcc_boot)
```

```{r}
biomarkers_table_list.df <- readxl::read_xlsx("../Data/Biomarkers_Payami_medRxiv.xlsx")
```



```{bash sparcc on PD, eval=FALSE}
cd ~/PayamiH/project_1_microGenPD/output/SparCC/

srun -c8 --mem=32G -pshort --pty bash

fastspar_bootstrap --otu_table input_tables/features_counts_PD.tsv --number 1000 --prefix input_tables/bootstrapped_tables/PD_

fastspar --iterations 100 --threads 8 --threshold 0.1 --seed 1234 --otu_table input_tables/features_counts_PD.tsv \--correlation SparCC_results/Species_SparCC_PD_Cor_Matrix.txt --covariance SparCC_results/Species_SparCC_PD_Cov_Matrix.txt


# calculate P-values in two steps

```


## Definition of subtypes based on Dirichlet Multinomial clustering

# Association between top PD associated taxa and genotypes

First challenge is defining top taxa. Haydeh chose based on signifincance 
theshoulds and correlation network of the taxa. The strategy is below




# R session

```{r}
sessionInfo()
```

