---
title: "PD, microbiome and host genetics"
author: "Giacomo Antonello, Zachary D. Wallen, Chad Murchinson, Levi Waldron, Haydeh Payami"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: hide
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
```

```{r, include=FALSE}
# Check and install necessary CRAN packages
CRAN_required_packages <- c("tidyverse", "gwasrapidd", "DT", "kableExtra", "data.table", "ggpubr", "BiocManager", "readxl","kableExtra")

install.packages(
  CRAN_required_packages[!(CRAN_required_packages %in% installed.packages())] 
)

# Check and install necessary Bioconductor packages
BioC_required_packages <- c("Maaslin2", "ANCOMBC", "biomaRt", "mia", "phyloseq", "microbiome","SIAMCAT", "fgsea")
BioC_required_packages <- BioC_required_packages[!(BioC_required_packages %in% installed.packages())]

if(length(BioC_required_packages) > 0) {
  BiocManager::install(BioC_required_packages[!(BioC_required_packages %in% installed.packages())],
                       ask = FALSE,
                       update = TRUE)
}


# load basic data manipulation packages
library(phyloseq)
library(DT)
library(kableExtra)
library(MultiAssayExperiment)
library(ggpubr)
library(tidyverse)

theme_set(theme_light())
# load previously wrangled data
load("../Data/wallenZD_data.mae.RData")

rm(BioC_required_packages,CRAN_required_packages)
gc(verbose = FALSE)
```

Below is the data sources for Wallen et al. 2022 (Nat Comms). This has been 
wrangled by Giacomo to make one single data object that would, hopefully, speed
things up in future analyses. This data object contains microbiome, pathway, 
participants data and genetic data.

```{r}
wallenZD_data.mae
```

Genetic data have been retrieved with the protocol available in `00_WallenZD_DataPreparation.html`.
In brief:

  1. The GWAS catalog was searched for variants associated with the ontology term
  `MONDO_0005180`, and downloaded. all this with 
  [gwasrapidd](https://rmagno.eu/gwasrapidd/articles/gwasrapidd.html). 
  Catalog's version is not stated, I'm assuming that since the packages uses the `REST API`, 
  it should be the newest.
  
  2. Gene annotation was taken from `biomaRt`, but the GWAS catalog had annotations
  too. Naming is based on the HUGO nomenclature.

# Assocation between PD markers and SNCA and HLA gene variants

## Cluster-wide biomarkers score (sum all)

Haydeh has manually curated a list of bacteria that go together into 6 different
clusters elevated or decreased in PD. To see the genetic impact of PD associated
variants on these bacteria without having to pay a high multiple testing price.

As a first simple approach we will sum all relative abundances together. Below 
is an example of the cluster names and bacteria associated with them.

```{r}
biomarkers_medrxiv.df <- readxl::read_xlsx("../Data/Biomarkers_Payami_medRxiv.xlsx")

datatable(biomarkers_medrxiv.df %>% select(-species_metaphlan))
```

For each sample, we can calculate the sum of all bacteria within each cluster as
one single value and use these as outcome variables. See below for the calculated
values.

```{r}

cluster_sums.df <- left_join(biomarkers_medrxiv.df, 
          as.data.frame(assay(wallenZD_data.mae, "metaphlan_RelAbund.se")/100) %>%
            rownames_to_column("species_metaphlan"), 
          by = "species_metaphlan") %>% 
  split.data.frame(f = .$cluster_name) %>% 
  lapply(select, starts_with("D")) %>% 
  lapply(colSums) %>% 
  bind_cols(sample_name = names(.[[1]])) %>% 
  relocate(sample_name)

cluster_sums.df %>% 
  mutate_if(is.numeric, round, 3) %>% 
datatable()
```

These variables are far from normal, and log transformation does not help.

```{r, fig.cap= "Distribution of cumulative sum of relative abundances of cluster member species."}
cluster_sums.df %>% 
  select(-sample_name) %>% 
  reshape2::melt() %>% 
  mutate(
    log2Transf = log2(value + 1),
    sq.root = sqrt(value)
    ) %>% 
  reshape2::melt(
    variable.name = "Transformation"
    ) %>% 
  mutate(
    Transformation = case_when(
      Transformation == "value" ~ 'crude',
      Transformation == "log2Transf" ~ 'log2',
      Transformation == "sq.root" ~ 'sq.root',
      TRUE ~ 'error'
      )
    ) %>% 
  ggplot(aes(x = value, color = Transformation)) + 
  geom_histogram(fill = "transparent") + 
  facet_wrap(~variable) 
```

# Genotype dosages x PD microbiome clusters (SNCA and HLA)

```{r}

samples_to_keep.df <- colData(wallenZD_data.mae[["metaphlan_RelAbund.se"]]) %>%
  as.data.frame() %>% 
  filter(
    collection_method != "swab",
    !is.na(Age_at_collection),
    !is.na(seqs_scaled),
    !is.na(Case_status)
    )

variants_to_keep_rowData.df <- as.data.frame(rowData(wallenZD_data.mae[["genotypes_dosages.se"]])[grepl("HLA|SNCA", rowData(wallenZD_data.mae[["genotypes_dosages.se"]])$GENE),])

per_sample_dosages.df <- t(assay(wallenZD_data.mae, "genotypes_dosages.se"))[,rownames(variants_to_keep_rowData.df)] %>% 
  as.data.frame() %>% 
  set_names(paste(variants_to_keep_rowData.df$CHROM, variants_to_keep_rowData.df$POS, variants_to_keep_rowData.df$REF, variants_to_keep_rowData.df$ALT, sep = "_")) %>% 
  rownames_to_column("sample_name")

data_alltogether.df <- inner_join(cluster_sums.df, per_sample_dosages.df, by = "sample_name") %>% 
  inner_join(samples_to_keep.df, by = "sample_name")

datatable(as_tibble(variants_to_keep_rowData.df) %>% 
            rename(`Imputation Quality (R2)` = R2))
```

Inclusion criteria:
  
  - Genetic variants were included if retained if:
    - they were associated with PD in any GWAS catalog entry
    - they were found in the Genetic data
    - they had a gene name associated to them
    The last step is important because it is how I defined that a variant *is* in 
    a gene or not
  - Collection method should not be `swab` 
  - Available data for age, sequencing depth, microbiome PD cumulative scores, genetics

This results in `r nrow(data_alltogether.df)` samples for the analysis, with variant's counts as follows:

```{r}
cbind(data_alltogether.df %>% 
  select(starts_with("chr")) %>% 
  mutate_all(round, 0) %>% 
  sapply(table) %>% 
  t(), 
  data_alltogether.df %>% 
    select(starts_with("chr")) %>% 
    mutate_all(round, 0) %>% 
    sapply(function(x) table(x) %>% prop.table()) %>% 
    t()) %>% 
  .[,c(1,4,2,5,3,6)] %>%   
  as.data.frame() %>% 
  set_names(make.unique(colnames(.))) %>% 
  rownames_to_column("id") %>%
  separate_wider_delim(cols = id, names = c("CHROM", "POS", "REF", "ALT"), delim = "_", too_few = "align_start") %>% 
  rename(
    `0 (N)` = `0`,
    `0 (%)` = `0.1`,
    `1 (N)` = `1`,
    `1 (%)` = `1.1`,
    `2 (N)` = `2`,
    `2 (%)` = `2.1`
  ) %>% 
  mutate_at(.vars = vars(c(`0 (%)`,`1 (%)`,`2 (%)`)), .funs = function(x) round(x*100, 2)) %>% 
  
  kbl(caption = "Number of samples with alternate allele dosages. NB: no distinction is done for 1|0 or 0|1, as both carry one alternate allese. Values are shown as total counts and percentages") %>% 
  kable_styling() %>% 
  add_header_above(c("Variant Information" = 4, "Distribution of dosages in the cohort studied" = 6)) %>% 
  column_spec(c(4,6,8), border_right = TRUE)
```


Now perform the model as follows, which allows the interaction term and will 
spit out coefficients for both variables alone.

$$
ClusterSum_i \sim Age + Seq.Depth + PD + Variant_k + PD \times Variant_k
$$

```{r bugs and genetics with lm}
clusters_vs_dosages.lm.list <- lapply(
  make.names(colnames(select(cluster_sums.df, -sample_name))) %>% set_names(colnames(select(cluster_sums.df, -sample_name))), 
  function(cl) {
    sapply(
      colnames(select(per_sample_dosages.df, -sample_name)), 
      function(gt) {
        f.la <- as.formula(
          paste(cl, "~ Age_at_collection + seqs_scaled + ", paste(gt, "Case_status", sep = "*"))
        )
        return(
          lm(f.la, data = data_alltogether.df %>% rename_all(.funs =make.names))
        )
      },
      USE.NAMES = TRUE,
  simplify = FALSE
    )
    
  }
)

# number of tests to correct for as bonferroni: number of cluster + variants tested
n_tests <- length(clusters_vs_dosages.lm.list) * 2
  
lm_results_corrected_bonferoni.df <- lapply(clusters_vs_dosages.lm.list, function(x) purrr::reduce(lapply(x, broom::tidy), rbind.data.frame)) %>% 
         bind_rows(.id = "cluster_name") %>% 
  mutate(
    p.adj_bonferroni = p.value * n_tests,
    p.adj_bonferroni = ifelse(p.adj_bonferroni >1, 1, p.adj_bonferroni),
    significant = ifelse(p.adj_bonferroni < 0.1, TRUE, FALSE)
    )

```

We correct p-values using the Bonferroni method accounting for 6 clusters tested
against 2 genes. This way, we are being a bit less stringent than necessary, but 
this can be fixed later. Significance was set to `Bonferroni p.adj < 0.1`. All
Cluster-variant associations discussed below are significant.

```{r bugs and genetics results table}
lm_results_corrected_bonferoni.df %>%
  filter(grepl("chr[0-9]", term), ) %>% 
  arrange(desc(abs(estimate))) %>% 
  select(-significant) %>% 
  mutate_if(is.numeric, round, 4) %>% 
  reactable::reactable(filterable = TRUE, searchable = TRUE, sortable = TRUE, columns = list(term = reactable::colDef(width = 250)))
```

Now some plots. For the sake of exploration, we will consider all associations
with crude `p-value < 0.05` and plot them.

```{r}
data_plot.tmp <- lm_results_corrected_bonferoni.df %>% 
  filter(grepl("chr", term)) %>% 
  separate_wider_delim(cols = term, names = c("variant", "PD"), delim = ":", too_few = "align_start") %>% 
  mutate(PD = case_when(
    PD == "Case_statusPD" ~ "PD",
    is.na(PD) ~ "Ctrl",
    TRUE ~ "Other"
  )) %>% 
  mutate(
    color_manual = case_when(
      p.value < 0.05 & PD == "Ctrl" ~ "skyblue3",
      p.value < 0.05 & PD == "PD" ~ "gold3",
      p.adj_bonferroni >= 0.05 ~ "gray"
  )
  )

ggplot() +
  geom_col(data = data_plot.tmp, mapping = aes(x = variant, y = estimate, fill = color_manual)) + 
  scale_fill_identity() +
  facet_wrap(~ cluster_name) +
  geom_hline(yintercept = 0, color = "gray30") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(
    y = "Beta estimates", 
    title = "Full Betas range"
  )

```

Only a few associations are significant here. Below you can find all significant
ones in boxplots taking into consideration genotypes and PD case/control status.
Plots are sorted by decreasing absolute $\beta$ values.

```{r, fig.height=12, fig.width=8}
plot_top_assoc <- lm_results_corrected_bonferoni.df %>%
  filter(grepl("chr[0-9]", term), p.value < 0.05) %>% 
  arrange(desc(abs(estimate)))

library(foreach)

ggarrange(plotlist = lapply(1:nrow(plot_top_assoc), function(i) {
  clst <- plot_top_assoc$cluster_name[i]
  varnt <- strsplit(plot_top_assoc$term[i], split = ":", fixed = TRUE) %>% sapply("[", 1)
  
  return(ggplot(data_alltogether.df %>% set_names(make.names(colnames(.))),
         aes(
           x = as.factor(round(eval(
             parse(text = varnt)
           ), 0)),
           y = eval(
             parse(text = make.names(clst)
           )),
           color = Case_status
         )) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2)) +
    geom_boxplot(alpha = 0.4,
                 show.legend = FALSE,
                 outliers = FALSE) +
    scale_color_manual(values = c("skyblue3", "gold3")) +
    labs(
      x = "Genotype Dosages (rounded)",
      y = "Clust.Score (crude)",
      title = paste(clst, "vs", varnt),
      color = "Disease status"
    )
  )
  }), ncol = 2, nrow = 4,
  common.legend = TRUE)
```

# Approach 2: Microbiome wide regression with another method, followed by GSEA (WIP)

As a sensitivity analysis, we can agnostically test all bacteria for association with 
genetic variants in interaction with PD. We can use `edgeR`, suggested by Levi 
after a recent benchmark as showing a nice overlap with expected signals based
on well characterized microbiome datasets. After testing, we can perform a gene 
set enrichment analysis using as gene sets:

  - the clusters generated by Haydeh
  - Other clusters as reported in `bugphyzz`, a repository of experimental data 
  on bacterial physiology. Second one is broader, so it could be an extention if
  we think it is necessary

## LinDA on SNCA and HLA

```{r}
samples_to_keep.df <- colData(wallenZD_data.mae[["metaphlan_RelAbund.se"]]) %>%
  as.data.frame() %>% 
  filter(collection_method != "swab")

variants_to_keep_rowData.df <- as.data.frame(rowData(wallenZD_data.mae[["genotypes_dosages.se"]])[grepl("HLA|SNCA", rowData(wallenZD_data.mae[["genotypes_dosages.se"]])$GENE),])

per_sample_dosages.df <- t(assay(wallenZD_data.mae, "genotypes_dosages.se"))[,rownames(variants_to_keep_rowData.df)] %>% 
  as.data.frame() %>% 
  set_names(paste(variants_to_keep_rowData.df$CHROM, variants_to_keep_rowData.df$POS, variants_to_keep_rowData.df$REF, variants_to_keep_rowData.df$ALT, sep = "_")) %>% 
  rownames_to_column("sample_name")

meta_with_variants <- inner_join(per_sample_dosages.df, samples_to_keep.df, by = "sample_name")

```

```{r}
if(!("LinDA" %in% installed.packages())){
  # using my forked LinDA method to simply allow better side effect handling (information output)
  devtools::install_github("zhouhj1994/LinDA")
  }

library(LinDA)

genotypes_vs_microbiome_linda.list <- sapply(grep("^chr", colnames(meta_with_variants), value = TRUE), 
       function(gt)
       suppressMessages(linda(
  otu.tab = assay(wallenZD_data.mae, "metaphlan_Counts.se")[,meta_with_variants$sample_name],
  meta = meta_with_variants,
  formula = paste("~ Age_at_collection + seqs_scaled + ", paste(gt, "Case_status", sep = "*")), 
  p.adj.method = "BH", 
  winsor.quan = 0.97, 
  alpha = 0.1,
  prev.cut = 0.05)),
  USE.NAMES = TRUE,
  simplify = FALSE
  )

linda_results_wrangled_variants_SNCA_HLA.df <- lapply(genotypes_vs_microbiome_linda.list, "[[", "output") %>% 
  lapply(function(x) x[grepl("chr", names(x))] %>% set_names(c("variantOnly", "interactionPD")) %>% lapply(rownames_to_column, "taxon")) %>% 
  lapply(
    bind_rows, .id = "covariate"
  ) %>% 
  bind_rows(.id = "variantName")

```


## GSEA on variant-only effects {.tabset}


**Interesting result**: `rs983361` follows an opposite trend as for fiber degraders 
enrichment (last tabset). Also, it is not significant in the previous analysis, 
but it is here. This suggests the value of trying an orthogonal approach as 
sensitivity analysis.

```{r}
# fgsea on variant only effects
variant_only_t_stats <- lapply(genotypes_vs_microbiome_linda.list, "[[", "output") %>% 
  lapply("[[", 3) %>% 
  lapply(function(x) set_names(x$stat, rownames(x)))

library(fgsea)

gsea_on_variantsonly <- lapply(variant_only_t_stats, function(stats) 
                               fgsea(
  pathways = split(biomarkers_medrxiv.df$species_metaphlan, biomarkers_medrxiv.df$cluster_name),
  stats = stats))

gsea_on_variantsonly_someSignif <- gsea_on_variantsonly[sapply(gsea_on_variantsonly, function(x) any(x$padj < 0.05))]
```


```{r, results="asis"}
for (i in names(gsea_on_variantsonly_someSignif)) {
    cat("\n###", i, sep = " ")
      
    cat("\n")
    
    variants_to_keep_rowData.df %>%
      filter(POS == strsplit(i, "_")[[1]][2]) %>%
      select(CHROM, POS, REF, ALT, RSID, GENE, R2) %>%
      kbl() %>%
      print()
    
    cat("\n")
    
    print(
      plotGseaTable(
      pathways = split(biomarkers_medrxiv.df$species_metaphlan, biomarkers_medrxiv.df$cluster_name), 
      stats = variant_only_t_stats[[i]],
      fgseaRes = gsea_on_variantsonly_someSignif[[i]])
      )
    cat("\n")
}
```

## {-}

## GSEA on variant-PD effects {.tabset}

```{r}
# fgsea on variant only effects
variant_PD_t_stat <- lapply(genotypes_vs_microbiome_linda.list, "[[", "output") %>% 
  lapply("[[", 5) %>% 
  lapply(function(x) set_names(x$stat, rownames(x)))

library(fgsea)

gsea_on_variant_PD <- lapply(variant_PD_t_stat, function(stats) 
                               fgsea(
  pathways = split(biomarkers_medrxiv.df$species_metaphlan, biomarkers_medrxiv.df$cluster_name),
  stats = stats))

gsea_on_variant_PD_someSignif <- gsea_on_variant_PD[sapply(gsea_on_variant_PD, function(x) any(x$padj < 0.05))]
```

```{r, results="asis"}
for (i in names(gsea_on_variant_PD_someSignif)) {
    cat("\n###", i, sep = " ")
      
    cat("\n")
    
    variants_to_keep_rowData.df %>%
      filter(POS == strsplit(i, "_")[[1]][2]) %>%
      select(CHROM, POS, REF, ALT, RSID, GENE, R2) %>%
      kbl() %>%
      print()
    
    cat("\n")
    
    print(
      plotGseaTable(
      pathways = split(biomarkers_medrxiv.df$species_metaphlan, biomarkers_medrxiv.df$cluster_name), 
      stats = variant_only_t_stats[[i]],
      fgseaRes = gsea_on_variant_PD_someSignif[[i]])
      )
    cat("\n")
}
```

```{r, eval=FALSE}
library(edgeR)
data_filt <- wallenZD_data.mae[["metaphlan_Counts.se"]][,colData(wallenZD_data.mae[["metaphlan_Counts.se"]])$collection_method != "swab"]
metadata.df <- colData(wallenZD_data.mae[["metaphlan_Counts.se"]]) %>% as.data.frame()
counts.mtx <- assay(wallenZD_data.mae, "metaphlan_Counts.se") %>% as.matrix()

glmFit.SummarizedExperiment(wallenZD_data.mae[["metaphlan_Counts.se"]], design = model.matrix())
```


# Open Questions

  - Do Fiber degraders correlate with fiber-rich diets?
  - Do Bifidos correlate with lactose rich diets?
  - Do Bifidos correlate with the variant in LCT gene? Do these people have it at 
  all in the VCFs?
  
# R session

```{r}
sessionInfo()
```

