---
title: "PD, microbiome and host genetics"
author: "Giacomo Antonello, Zachary D. Wallen, Chad Murchinson, Levi Waldron, Haydeh Payami"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: hide
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
```

```{r, include=FALSE}
# Check and install necessary CRAN packages
CRAN_required_packages <- c("tidyverse", "gwasrapidd", "DT", "kableExtra", "data.table", "ggpubr", "BiocManager", "readxl","kableExtra")

install.packages(
  CRAN_required_packages[!(CRAN_required_packages %in% installed.packages())] 
)

# Check and install necessary Bioconductor packages
BioC_required_packages <- c("Maaslin2", "ANCOMBC", "biomaRt", "mia", "phyloseq", "microbiome","SIAMCAT")
BioC_required_packages <- BioC_required_packages[!(BioC_required_packages %in% installed.packages())]

if(length(BioC_required_packages) > 0) {
  BiocManager::install(BioC_required_packages[!(BioC_required_packages %in% installed.packages())],
                       ask = FALSE,
                       update = TRUE)
}


# load basic data manipulation packages
library(phyloseq)
library(DT)
library(kableExtra)
library(MultiAssayExperiment)
library(tidyverse)
# load previously wrangled data
load("../Data/wallenZD_data.mae.RData")

rm(BioC_required_packages,CRAN_required_packages)
gc(verbose = FALSE)
```

```{r}
wallenZD_data.mae
```

# Assocation between PD markers and PD genetics

## Cluster-wide biomarkers score (sum all)

Haydeh has manually curated a list of bacteria that go together into 6 different
clusters elevated or decreased in PD. To see the genetic impact of PD associated
variants on these bacteria without having to pay a high multiple testing price.

As a first simple approach we will sum all relative abundances together. Below 
is an example of the cluster names and bacteria associated with them.

```{r}
biomarkers_medrxiv.df <- readxl::read_xlsx("../Data/Biomarkers_Payami_medRxiv.xlsx")

datatable(biomarkers_medrxiv.df %>% select(-species_metaphlan))
```

For each sample, we can calculate the sum of all bacteria within each cluster as
one single value and use these as outcome variables. See below for the calculated
values.

```{r}

cluster_sums.df <- left_join(biomarkers_medrxiv.df, 
          as.data.frame(assay(wallenZD_data.mae, "metaphlan_RelAbund.se")/100) %>%
            rownames_to_column("species_metaphlan"), 
          by = "species_metaphlan") %>% 
  split.data.frame(f = .$cluster_name) %>% 
  lapply(select, starts_with("D")) %>% 
  lapply(colSums) %>% 
  bind_cols(sample_name = names(.[[1]])) %>% 
  relocate(sample_name)

cluster_sums.df %>% 
  mutate_if(is.numeric, round, 3) %>% 
datatable()
```

These variables are far from normal, and log transformation does not help. This 
is due to 

```{r}
cluster_sums.df %>% 
  select(-sample_name) %>% 
  reshape2::melt() %>% 
  mutate(
    log2Transf = log2(value + 1),
    sq.root = sqrt(value)
    ) %>% 
  reshape2::melt(
    variable.name = "Transformation"
    ) %>% 
  mutate(
    Transformation = case_when(
      Transformation == "value" ~ 'crude',
      Transformation == "log2Transf" ~ 'log2',
      Transformation == "sq.root" ~ 'sq.root',
      TRUE ~ 'error'
      )
    ) %>% 
  ggplot(aes(x = value, color = Transformation)) + 
  geom_histogram(fill = "transparent") + 
  facet_wrap(~variable) + 
  theme_light()
```

## Genotype dosages x PD microbiome clusters (lm)

Inclusion-exclusion criteria:
  
  - Collection method should not be `swab` 
  - Variants of interest should be mapped to `HLA` and `SNCA` genes
  - samples should be in common between genetics and microbiome

```{r}

samples_to_keep.df <- colData(wallenZD_data.mae[["metaphlan_RelAbund.se"]]) %>%
  as.data.frame() %>% 
  filter(collection_method != "swab")

variants_to_keep_rowData.df <- as.data.frame(rowData(wallenZD_data.mae[["genotypes_dosages.se"]])[grepl("HLA|SNCA", rowData(wallenZD_data.mae[["genotypes_dosages.se"]])$GENE),])

per_sample_dosages.df <- t(assay(wallenZD_data.mae, "genotypes_dosages.se"))[,rownames(variants_to_keep_rowData.df)] %>% 
  as.data.frame() %>% 
  set_names(paste(variants_to_keep_rowData.df$CHROM, variants_to_keep_rowData.df$POS, sep = "_")) %>% 
  rownames_to_column("sample_name")

data_alltogether.df <- inner_join(cluster_sums.df, per_sample_dosages.df, by = "sample_name") %>% 
  inner_join(samples_to_keep.df, by = "sample_name")

```

Now perform the model as follows, which allows the interaction term and will 
spit out coefficients for both variables alone.

$$
ClusterSum_i \sim Age + Seq.Depth + PD \times Variant_k
$$



```{r bugs and genetics with lm}
clusters_vs_dosages.lm.list <- lapply(
  make.names(colnames(select(cluster_sums.df, -sample_name))) %>% set_names(colnames(select(cluster_sums.df, -sample_name))), 
  function(cl) {
    sapply(
      colnames(select(per_sample_dosages.df, -sample_name)), 
      function(gt) {
        f.la <- as.formula(
          paste(cl, "~ Age_at_collection + seqs_scaled + ", paste("Case_status", gt, sep = "*"))
        )
        return(
          lm(f.la, data = data_alltogether.df %>% rename_all(.funs =make.names))
        )
      },
      USE.NAMES = TRUE,
  simplify = FALSE
    )
    
  }
)

# number of tests to correct for as bonferroni: number of cluster + variants tested
n_tests <- length(clusters_vs_dosages.lm.list) + max(sapply(clusters_vs_dosages.lm.list, length))
  
lm_results_corrected_bonferoni.df <- lapply(clusters_vs_dosages.lm.list, function(x) purrr::reduce(lapply(x, broom::tidy), rbind.data.frame)) %>% 
         bind_rows(.id = "cluster_name") %>% 
  mutate(p.adj_bonferroni = p.value / n_tests)

lm_results_corrected_bonferoni_signif.df <- lm_results_corrected_bonferoni.df %>% 
  filter(p.adj_bonferroni < 0.1, grepl("chr[0-9]", term)) %>% 
  mutate_if(is.numeric, round, 3)
```

We adjust for False discovery rate with Bonferroni correction, dividing 
P-values by the number of clusters and the number of variants tested (CorrValue = `r print(n_tests)`).

Several associations are found with PD and with variants (or their interaction).
Interestingly, there is a strong association between **Fiber Degraders** and variants
in the SNCA gene. Below is a table of significant results (`Bonferroni Q-value < 0.1`) sorted by decreasing absolute $\beta$ coefficient.

```{r bugs and genetics results table}
lm_results_corrected_bonferoni_signif.df %>%
  arrange(desc(abs(estimate))) %>% 
  datatable()
```

This variant, `chr4:89840793`, is particularly interesting because Fiber degraders
are negatively associated in case of an interaction between that and PD, but 
they are increased in case of carrying more alternative alleles. The distribution
is:

```{r}
ggplot(data_alltogether.df, aes(x = as.factor(round(chr4_89840793, 0)), y = `Fiber Degraders`, color = Case_status)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2)) +
  geom_boxplot(alpha = 0.3) +
  theme_light() +
  labs(
    x = "Genotype Dosages (rounded)",
    title = "Chr4:89840793 vs Fiber degraders"
  )
```


```{r}
lm_results_corrected_bonferoni.df %>% 
  filter(grepl("^chr", term)) %>% 
  ggplot(aes(x = term, y = cluster_name, fill = estimate)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    x = "SNCA and HLA variants",
    y = "Microbiome cluster in PD (Haydeh)",
    fill = "Beta (lm)"
  )
```


```{r}
lm_results_corrected_bonferoni.df %>% 
  filter(grepl("PD:chr", term)) %>% 
  ggplot(aes(x = term, y = cluster_name, fill = estimate)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    x = "SNCA and HLA variants",
    y = "Microbiome cluster in PD (Haydeh)",
    fill = "Beta (lm)"
  )


```

Modeling assumptions are far from met, but we find a few associations with variants
alone and variants in interaction with these bacterial clusters.

Questions are the following:

  - Bacteria associated with variants alone 
  - Bacteria associated with a variant in case of PD but not in case of controls

# BugSigDbR package to compare pre-published bugs

```{r}

```



# R session

```{r}
sessionInfo()
```

