---
title: "PD, microbiome and host genetics"
author: "Giacomo Antonello, Zachary D. Wallen, Chad Murchinson, Levi Waldron, Haydeh Payami"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: hide
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
```

```{r}
# Check and install necessary CRAN packages
CRAN_required_packages <- c("tidyverse", "gwasrapidd", "DT", "kableExtra", "data.table", "ggpubr", "BiocManager", "readxl","kableExtra")

install.packages(
  CRAN_required_packages[!(CRAN_required_packages %in% installed.packages())] 
)

# Check and install necessary Bioconductor packages
BioC_required_packages <- c("Maaslin2", "ANCOMBC", "biomaRt", "mia", "phyloseq", "microbiome","SIAMCAT")

BiocManager::install(
  BioC_required_packages[!(BioC_required_packages %in% installed.packages())], ask = FALSE, update = TRUE 
)

# load basic data manipulation packagesß
library(phyloseq)
library(DT)
library(kableExtra)
library(MultiAssayExperiment)
library(tidyverse)
# load previously wrangled data
load("../Data/wallenZD_data.mae.RData")
```

# Assocation between PD markers and PD genetics

## Table of biomarkers

```{r}
biomarkers_medrxiv.df <- readxl::read_xlsx("../Data/Biomarkers_Payami_medRxiv.xlsx")

datatable(biomarkers_medrxiv.df %>% select(-original_taxon_name))
```


Step 1, get list of genetic variants found in WallenZD under the `SNCA` and 
`HLA` gene annotations.

```{r}
# remove samples that were collected with swabs instead of the correct kit
samples_no_swab <- colData(wallenZD_data.mae)$sample_name[colData(wallenZD_data.mae)$collection_method != "swab"]
#
variants_SNCA_HLA <- rownames(rowData(wallenZD_data.mae[["genotypes.se"]])[grepl("HLA|SNCA", rowData(wallenZD_data.mae[["genotypes.se"]])$GENE),])

bugs_signatures <- biomarkers_medrxiv.df$original_taxon_name

genotypes_with_metadata <- assay(wallenZD_data.mae, "genotypes.se")[variants_SNCA_HLA,intersect(samples_no_swab, colnames(assay(wallenZD_data.mae, "genotypes.se")))] %>% 
  t() %>% 
  as.data.frame() %>%
  mutate_all(.funs = function(x) case_when(
    x == "0|0" ~ "HomRef",
    x == "1|1" ~ "HomAlt",
    TRUE ~ "Het"
  ) %>% factor(levels = c("HomRef", "Het", "HomAlt"))) %>% 
  rownames_to_column("sample_name") %>% 
  inner_join(as.data.frame(colData(wallenZD_data.mae)), by = "sample_name") %>% 
  filter(sample_name %in%  samples_no_swab) 

rownames(genotypes_with_metadata) <-  genotypes_with_metadata$sample_name
```

Step 2, perform the model, with the following formula

Group microbiome features together by summing their abundances, followed by log
transformation.

$$
bug \sim 
$$

Step 3, perform additive associations

```{r bugs and genetics with Maaslin2}

microbes_matrix <- as.matrix(assay(wallenZD_data.mae, "metaphlan_RelAbund.se"))[bugs_signatures,genotypes_with_metadata$sample_name] / 100

dir.create("output/genetics_vs_microbes_PD/Maaslin2", recursive = TRUE, showWarnings = FALSE)

maaslin_per_variant <- lapply(variants_SNCA_HLA,
       function(rsid)
         Maaslin2::Maaslin2(
           input_data = t(microbes_matrix),
           input_metadata = genotypes_with_metadata,
           output = file.path("output/genetics_vs_microbes_PD/Maaslin2",rsid),
  fixed_effects = c(rsid, "Case_status", paste("Case_status", rsid, sep = "*"), "seqs_scaled", "Age_at_collection"),
  # min_prevalence = 0.05,
  normalization = 'NONE',
  max_significance = 0.05,
  standardize = FALSE,
  plot_heatmap = FALSE,
  plot_scatter = FALSE,
  heatmap_first_n = 0
  )
)

rsID_associations_bimarker_bugs_PD.list <- lapply(maaslin_per_variant, "[[", "results") %>%
  lapply(filter, grepl("^rs", metadata))

```

```{r bugs and genetics with ANCOMBC}
ps.counts_genetics <- phyloseq(
        otu_table(
          as.matrix(assay(wallenZD_data.mae, "metaphlan_Counts.se"))[bugs_signatures,genotypes_with_metadata$sample_name] / 100, taxa_are_rows = TRUE),
        tax_table(as.matrix(rowData(wallenZD_data.mae[["metaphlan_Counts.se"]]))[bugs_signatures,]),
        sample_data(genotypes_with_metadata)
      )

dir.create("output/genetics_vs_microbes_PD/ANCOMBC", recursive = TRUE, showWarnings = FALSE)

for(rsid in variants_SNCA_HLA){
  tmp_res <- tryCatch(
    ANCOMBC::ancombc(
      data = ps.counts_genetics,
      prv_cut = 0.05,
      formula = paste(
        c(
          rsid,
          "Case_status",
          paste("Case_status", rsid, sep = "*"),
          "seqs_scaled",
          "Age_at_collection"
        ),
        collapse = " + "
      ),
      p_adj_method = "BH",
      verbose = FALSE,
      alpha = 0.1,
      n_cl = 4 
    ),
    error = function(e)
      return(e)
  )
  saveRDS(
    tmp_res,
    paste0(
      "output/genetics_vs_microbes_PD/ANCOMBC/ancombc_results_",
      rsid,
      "_raw.Rds"
    )
  )
  
}
```

```{r summary diff abund maaslin2}



```

See results for `SNCA` genes, just effect of variant

```{r summary results SNCA genes}
snca_variants.df <- rowData(wallenZD_data.mae[["genotypes.se"]]) %>% 
  as.data.frame %>% 
  filter(grepl("SNCA", GENE))

datatable(snca_variants.df)
```

```{r}
ANCOMBC_raw_results <- list.files("output/genetics_vs_microbes_PD/ANCOMBC", full.names = TRUE) %>%
  set_names(str_extract(base::basename(.), "rs[0-9]*")) %>% 
  lapply(readRDS) %>% 
  # remove ancombc results that failed in the step above
  .[!grepl("error", sapply(., class))] %>% 
  lapply("[[", "res") %>% 
  purrr::transpose()


just_rs_lfc <- lapply(ANCOMBC_raw_results$lfc, function(x) x[,c(1,3,4)]) %>% 
  lapply(set_names, c("taxon", "Het", "Alt")) %>% 
  bind_rows(.id = "rsID")

head(just_rs_lfc)

ggplot(just_rs_lfc %>% pivot_longer(names_to = "Genotype", values_to = "lfc", cols = c("Het", "Alt")) %>% mutate(Genotype = factor(Genotype, levels = c("Het", "Alt"))), 
       aes(x = rsID, y = lfc, fill = Genotype)) +
  geom_col(position = position_dodge())+
  scale_color_manual(values = c("gray", "black")) +
  facet_wrap(~taxon) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    title = "SNCA variants"
  )

```


```{r summary diff abund ancom}
ANCOMBC_raw_results <- list.files("output/genetics_vs_microbes_PD/ANCOMBC", full.names = TRUE) %>%
  set_names(str_extract(base::basename(.), "rs[0-9]*")) %>% 
  lapply(readRDS) %>% 
  # remove ancombc results that failed in the step above
  .[!grepl("error", sapply(., class))] %>% 
  lapply("[[", "res") %>% 
  # extract results
  lapply(function(df) lapply(df, select, !contains(c("seqs_scaled","Age_at_collection")))) #%>% 
  lapply(bind_cols) #%>% 
  rename_all(.funs = function(x) str_remove(x, "\\.Case_statusPD")) %>% 
  mutate(
    taxon = ancombc_results_raw$res$lfc$taxon,
    FC = exp(lfc)
    ) %>% 
  relocate(taxon, lfc, FC) %>% 
  arrange(taxon)
  

list.files("output/genetics_vs_microbes_PD/ANCOMBC", full.names = TRUE) 

```

# alternative to clusters for biomarkers, variable importance in machine learning

```{r}
library(siamcat)
```

### R reimplementation of sparcc

SparCC is problematic on the Cheaha server, somehow the permutations step won't 
work.

```{r}
devtools::install_github("zdk123/SpiecEasi")

library(SpiecEasi)

PD_sparcc_base <- sparcc(
  assay(wallenZD_data.mae, "metaphlan_Counts.se") %>% 
  .[rowSums(.) > 0,PD_samples] %>% 
  rownames_to_column("OTU_id") %>% 
  filter(OTU_id != "UNKNOWN") %>% 
    column_to_rownames("OTU_id"), 
  iter = 100, th = 0.2)

PD_sparcc_boot <- sparccboot(data = assay(wallenZD_data.mae, "metaphlan_Counts.se") %>% 
  .[rowSums(.) > 0,PD_samples] %>% 
  rownames_to_column("OTU_id") %>% 
  filter(OTU_id != "UNKNOWN") %>% 
    column_to_rownames("OTU_id"), R = 1000)

saveRDS(PD_sparcc_boot, "PD_sparcc_boot.Rds")
#sparcc_pvalues <- pval.sparccboot(PD_sparcc_boot)
```

```{r}
biomarkers_table_list.df <- readxl::read_xlsx("../Data/Biomarkers_Payami_medRxiv.xlsx")

datatable(biomarkers_table_list.df)
```

## command line SparCC approach (not working anymore)

```{r, eval=FALSE}

control_samples <- colData(wallenZD_data.mae)$sample_name[colData(wallenZD_data.mae)$Case_status == "Control"]
PD_samples <- colData(wallenZD_data.mae)$sample_name[colData(wallenZD_data.mae)$Case_status == "PD"]

dir.create("output/SparCC/input_tables/bootstrapped_tables", recursive = TRUE, showWarnings = FALSE)
dir.create("output/SparCC/SparCC_results", recursive = TRUE, showWarnings = FALSE)

# write features table without all-zero columns formatted as SparCC likes them

assay(wallenZD_data.mae, "metaphlan_Counts.se") %>% 
  .[rowSums(.) > 0,PD_samples] %>% 
  rownames_to_column("OTU_id") %>% 
  filter(OTU_id != "UNKNOWN") %>% 
  write_tsv(file = "output/SparCC/input_tables/features_counts_PD.tsv")

assay(wallenZD_data.mae, "metaphlan_Counts.se") %>% 
  .[rowSums(.) > 0,control_samples] %>% 
  rownames_to_column("OTU_id") %>% 
  filter(OTU_id != "UNKNOWN") %>% 
  write_tsv(file = "output/SparCC/input_tables/features_counts_HC.tsv")


```

```{bash, eval=FALSE}
# mamba gave issues with setting up this particular environment, so I will need to use conda
module add Anaconda3

#installing without the following step results in an error:

conda config --set channel_priority disabled

conda env create -f /data/project/payami-lab/payamilab/fastspar.yaml

conda activate fastspar
```


```{bash sparcc on PD, eval=FALSE}
cd ~/PayamiH/project_1_microGenPD/output/SparCC/

srun -c8 --mem=32G -pshort --pty bash

fastspar --iterations 100 --threads 10 --threshold 0.1 --seed 1234 --otu_table input_tables/features_counts_PD.tsv --correlation SparCC_results/Species_SparCC_PD_Cor_Matrix.txt --covariance SparCC_results/Species_SparCC_PD_Cov_Matrix.txt

fastspar_bootstrap --otu_table input_tables/features_counts_PD.tsv --number 1000 --prefix input_tables/bootstrapped_tables/PD_

fastspar --iterations 100 --threads 8 --threshold 0.1 --seed 1234 --otu_table input_tables/features_counts_PD.tsv \--correlation SparCC_results/Species_SparCC_PD_Cor_Matrix.txt --covariance SparCC_results/Species_SparCC_PD_Cov_Matrix.txt


# calculate P-values in two steps

```


## Definition of subtypes based on Dirichlet Multinomial clustering

# Association between top PD associated taxa and genotypes

First challenge is defining top taxa. Haydeh chose based on signifincance 
theshoulds and correlation network of the taxa. The strategy is below




# R session

```{r}
sessionInfo()
```

