---
title: "Multi-dataset replicability of genetics vs PD associations"
author: "Giacomo Antonello, Zachary D. Wallen, Charles F. Murchinson, Levi Waldron, Haydeh Payami"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: hide
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
---

```{r knitr preferences, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
```

```{r library load}
library(tidyverse)
library(reactable)

```

# Interaction `variant x PD`

## Approach 1 -  Clusters abundance

```{r}
approach1_interaction_raw_results <- list(
  DoD = read_tsv("output_DoD/Q1/01_lm_interaction_Variant.x.PD.tsv") %>% mutate(Dataset = "DoD"),
  NGRC = read_tsv("output_NGRC/Q1/01_lm_interaction_Variant.x.PD.tsv") %>% mutate(Dataset = "NGRC"),
  UAB = read_tsv("output_UAB/Q1/01_lm_interaction_Variant.x.PD.tsv") %>% mutate(Dataset = "UAB")
) %>% 
  lapply(relocate, Dataset) %>% 
  lapply(filter, grepl(":", term))
```

### Significance based on Bonferroni-adjusted P-values

No interaction term was ever significant accounting for all independent tests
as defined before ($N_{clusters} \times N_{indep.Variants}$).

```{r}
tmp <- lapply(approach1_interaction_raw_results, filter, p.adj_bonferroni < 0.1) %>% 
  lapply(mutate_if, is.numeric, round, 4) %>% 
  lapply(reactable, sortable = TRUE, filterable = TRUE)
```

```{r}
tmp$DoD
```

```{r}
tmp$NGRC
```

```{r}
tmp$UAB
```

### Conclusions of approach 1 (ClusterSum)

  * Within dataset:
    - There were no Bonferroni-adjusted interaction associations
  * Across datasets:
    - No overlap of the few associations marginally significant considering raw
    p-values.
    
## Approach 2 - Regression of each microbe, followed by Microbe Set Enrichment analysis

### LinDA association only

```{r}
approach2_interaction_linda <- list(
  DoD = read_tsv("output_DoD/Q1/02_LinDA_regression_dosage.x.PD_FullResults.tsv") %>% mutate(Dataset = "DoD"),
  NGRC = read_tsv("output_NGRC/Q1/02_LinDA_regression_dosage.x.PD_FullResults.tsv") %>% mutate(Dataset = "NGRC"),
  UAB = read_tsv("output_UAB/Q1/02_LinDA_regression_dosage.x.PD_FullResults.tsv") %>% mutate(Dataset = "UAB")
) %>% 
  lapply(relocate, Dataset) %>% 
  lapply(filter, grepl("interactionPD", covariate)) 

```

```{r, results="asis"}

tables_as_list <- approach2_interaction_linda %>% 
  lapply(filter, padj < 0.1) %>% 
  lapply(mutate_if, is.numeric, round, 4) %>% 
  lapply(select, -lfcSE, -reject, -df, -covariate) %>% 
  lapply(reactable, sortable = TRUE, filterable = TRUE)


```

```{r}
tables_as_list$DoD
```

```{r}
tables_as_list$NGRC
```

```{r}
tables_as_list$UAB
```

Are there cases where the same variant:PD interaction term is significantly 
associated with the same bacterium in more than one dataset?

```{r}
tmp <- approach2_interaction_linda %>% 
  lapply(filter, padj < 0.1) %>% 
  reduce(full_join, by = c("taxon", "variantName"))

associations_replic <- tmp[rowSums(is.na(select(tmp, contains("log2FoldChange")))) < 2,]

linda_intersection_also_nonsignif <- approach2_interaction_linda %>% 
  lapply(filter, variantName %in% associations_replic$variantName, taxon %in% associations_replic$taxon) %>% 
  reduce(rbind.data.frame) %>% 
  mutate(alpha = ifelse(padj < 0.1, 1, 0.5))

reactable(linda_intersection_also_nonsignif)
```

There are not replications of associations across datasets, but multiple ones 
in each separately.

### Microbe Set Enrichment Analysis with LinDA `statistic`s

```{r}
approach2_linda_MSEA <- list(
  DoD = read_tsv("output_DoD/Q1/03_full_MSEA_results_interaction.tsv") %>% mutate(Dataset = "DoD"),
  NGRC = read_tsv("output_NGRC/Q1/03_full_MSEA_results_interaction.tsv") %>% mutate(Dataset = "NGRC"),
  UAB = read_tsv("output_UAB/Q1/03_full_MSEA_results_interaction.tsv") %>% mutate(Dataset = "UAB")
) %>% 
  lapply(relocate, Dataset)
```

Below is the table of union of Microbe sets enriched or depleted across
the three datasets.

```{r}
approach2_linda_MSEA  %>%  
  lapply(filter, padj < 0.1) %>% 
  lapply(select, -leadingEdge) %>% 
  reduce(full_join, by = c("clusterName", "snp_variant")) %>% 
  mutate_if(is.numeric, round, 4) %>% 
  reactable()
```

To better view the degree of replication, let's plot all and show where
FDR q-values are less than 0.1.

```{r, fig.caption = "Plot of variants with at least 2 datasets associated in MSEA using linda t-statistics on the interaction term. Semi-transparent bars indicate datasets with non-significant enrichment or depletion for that specific bacterial cluster and variant. Dotted lines indicate an arbitrary threshold of microbe set enrichment at |NES| > 1.2"}
tmp <- approach2_linda_MSEA %>% 
  lapply(filter, padj < 0.1) %>% 
  reduce(full_join, by = c("clusterName", "snp_variant"))

# restrict to rows that have at least 2 datasets replicated
# that is the row must contain between 0 and 1 NAs
associations_replic <- tmp[rowSums(is.na(select(tmp, contains("NES")))) < 2,]


MSEA_intersection_also_nonsignif <- approach2_linda_MSEA %>% 
  lapply(filter, snp_variant %in% associations_replic$snp_variant, clusterName %in% associations_replic$clusterName) %>% 
  reduce(rbind.data.frame) %>% 
  mutate(alpha = ifelse(padj < 0.1, 1, 0.5))
  
NES_barplot <- ggplot(MSEA_intersection_also_nonsignif, aes(x = clusterName, y = NES, fill = Dataset)) + 
  geom_col(position = position_dodge(), mapping = aes(alpha = alpha)) +
facet_wrap(~snp_variant) + 
  theme_light() + 
  geom_hline(yintercept = 1.2, col = "black", lty = "dotted") +
  geom_hline(yintercept = -1.2, col = "black", lty = "dotted") + 
  guides(alpha = "none") + 
  scale_fill_manual(values = ggsci::pal_jco()(4)[c(4,1,2)])

NES_barplot
```

### Conclusions from approach 2 (MSEA)

  * Four Variants show a significant enrichment of Opportunistic pathogens in 
  NGRC and UAB but not in DoD. 
  
  * Variant bottom right, which showed an interesting association with Fiber degraders 
  as `clusterSum` exclusively in DoD, shows a normalized depletion even stronger
  in NGRC, which would have been overlooked.
  
  * All in all, dataset specificity is still driving a large portion of the 
  microbiome variability, more than PD interaction with genetic variants can 
  capture with such small sample size.
  
  * Of note, while single interactions are maybe weak, we should consider additive
  effects of multiple variants in PD patients.
  
# Variant effect on PD

## Approach 1 -  Clusters abundance

```{r}
approach1_interaction_raw_results <- list(
  DoD = read_tsv("output_DoD/Q2/01_lm_dosage_in_PD.tsv") %>% mutate(Dataset = "DoD"),
  NGRC = read_tsv("output_NGRC/Q2/01_lm_dosage_in_PD.tsv") %>% mutate(Dataset = "NGRC"),
  UAB = read_tsv("output_UAB/Q2/01_lm_dosage_in_PD.tsv") %>% mutate(Dataset = "UAB")
) %>% 
  lapply(relocate, Dataset) %>% 
  lapply(filter, grepl(":", term))
```

### Significance based on Bonferroni-adjusted P-values

No interaction term was ever significant accounting for all independent tests
as defined before ($N_{clusters} \times N_{indep.Variants}$).

```{r}
tmp <- lapply(approach1_interaction_raw_results, filter, p.adj_bonferroni < 0.1) %>% 
  lapply(mutate_if, is.numeric, round, 4) %>% 
  lapply(reactable, sortable = TRUE, filterable = TRUE)
```

```{r}
tmp$DoD
```

```{r}
tmp$NGRC
```

```{r}
tmp$UAB
```

### Conclusions of approach 1 (ClusterSum)

  * Within dataset:
    - There were no variant associations with ClusterSums in each dataset,
    in any of the two significance thresholds chosen
  * Across datasets:
    - There were not associations within dataset that could be discussed
    
## Approach 2 - Regression of each microbe, followed by Microbe Set Enrichment analysis

### LinDA association only

```{r}
approach2_interaction_linda <- list(
  DoD = read_tsv("output_DoD/Q2/02_LinDA_regression_dosage_in_PD.tsv") %>% mutate(Dataset = "DoD"),
  NGRC = read_tsv("output_NGRC/Q2/02_LinDA_regression_dosage_in_PD.tsv") %>% mutate(Dataset = "NGRC"),
  UAB = read_tsv("output_UAB/Q2/02_LinDA_regression_dosage_in_PD.tsv") %>% mutate(Dataset = "UAB")
) %>% 
  lapply(relocate, Dataset) %>% 
  lapply(filter, grepl("interactionPD", covariate)) 

```

```{r, results="asis"}

tables_as_list <- approach2_interaction_linda %>% 
  lapply(filter, padj < 0.1) %>% 
  lapply(mutate_if, is.numeric, round, 4) %>% 
  lapply(select, -lfcSE, -reject, -df, -covariate) %>% 
  lapply(reactable, sortable = TRUE, filterable = TRUE)

```

```{r}
tables_as_list$DoD
```

```{r}
tables_as_list$NGRC
```

```{r}
tables_as_list$UAB
```

Are there cases where the same variant association term in PD is significantly 
associated with the same bacterium in more than one dataset?

```{r}
tmp <- approach2_interaction_linda %>% 
  lapply(filter, padj < 0.1) %>% 
  reduce(full_join, by = c("taxon", "variantName"))

associations_replic <- tmp[rowSums(is.na(select(tmp, contains("log2FoldChange")))) < 2,]

linda_intersection_also_nonsignif <- approach2_interaction_linda %>% 
  lapply(filter, variantName %in% associations_replic$variantName, taxon %in% associations_replic$taxon) %>% 
  reduce(rbind.data.frame) %>% 
  mutate(alpha = ifelse(padj < 0.1, 1, 0.5))

reactable(linda_intersection_also_nonsignif)
```

There are not replications of associations across datasets, but multiple ones 
in each separately.

### Microbe Set Enrichment Analysis with LinDA `statistic`s

```{r}
approach2_linda_MSEA <- list(
  DoD = read_tsv("output_DoD/Q2/03_full_MSEA_results_variant_in_PD.tsv") %>% mutate(Dataset = "DoD"),
  NGRC = read_tsv("output_NGRC/Q2/03_full_MSEA_results_variant_in_PD.tsv") %>% mutate(Dataset = "NGRC"),
  UAB = read_tsv("output_UAB/Q2/03_full_MSEA_results_variant_in_PD.tsv") %>% mutate(Dataset = "UAB")
) %>% 
  lapply(relocate, Dataset)
```

Below is the table of union of Microbe sets enriched or depleted across
the three datasets.

```{r}
approach2_linda_MSEA  %>%  
  lapply(filter, padj < 0.1) %>% 
  lapply(select, -leadingEdge) %>% 
  reduce(full_join, by = c("clusterName", "snp_variant")) %>% 
  mutate_if(is.numeric, round, 4) %>% 
  reactable()
```

To better view the degree of replication, let's plot all and show where
FDR q-values are less than 0.1.

```{r, fig.caption = "Plot of variants with at least 2 datasets associated in MSEA using linda t-statistics on the interaction term. Semi-transparent bars indicate datasets with non-significant enrichment or depletion for that specific bacterial cluster and variant. Dotted lines indicate an arbitrary threshold of microbe set enrichment at |NES| > 1.2"}
tmp <- approach2_linda_MSEA %>% 
  lapply(filter, padj < 0.1) %>% 
  reduce(full_join, by = c("clusterName", "snp_variant"))

# restrict to rows that have at least 2 datasets replicated
# that is the row must contain between 0 and 1 NAs
associations_replic <- tmp[rowSums(is.na(select(tmp, contains("NES")))) < 2,]


MSEA_intersection_also_nonsignif <- approach2_linda_MSEA %>% 
  lapply(filter, snp_variant %in% associations_replic$snp_variant, clusterName %in% associations_replic$clusterName) %>% 
  reduce(rbind.data.frame) %>% 
  mutate(alpha = ifelse(padj < 0.1, 1, 0.5))
  
NES_barplot <- ggplot(MSEA_intersection_also_nonsignif, aes(x = clusterName, y = NES, fill = Dataset)) + 
  geom_col(position = position_dodge(), mapping = aes(alpha = alpha)) +
facet_wrap(~snp_variant) + 
  theme_light() + 
  geom_hline(yintercept = 1.2, col = "black", lty = "dotted") +
  geom_hline(yintercept = -1.2, col = "black", lty = "dotted") + 
  guides(alpha = "none") + 
  scale_fill_manual(values = ggsci::pal_jco()(4)[c(4,1,2)])

NES_barplot
```


# Variant effect on Controls (NHC)

## Approach 1 -  Clusters abundance

```{r}
approach1_interaction_raw_results <- list(
  DoD = read_tsv("output_DoD/Q3/01_lm_dosage_in_Controls.tsv") %>% mutate(Dataset = "DoD"),
  NGRC = read_tsv("output_NGRC/Q3/01_lm_dosage_in_Controls.tsv") %>% mutate(Dataset = "NGRC"),
  UAB = read_tsv("output_UAB/Q3/01_lm_dosage_in_Controls.tsv") %>% mutate(Dataset = "UAB")
) %>% 
  lapply(relocate, Dataset) %>% 
  lapply(filter, grepl(":", term))
```

### Significance based on Bonferroni-adjusted P-values

No interaction term was ever significant accounting for all independent tests
as defined before ($N_{clusters} \times N_{indep.Variants}$).

```{r}
tmp <- lapply(approach1_interaction_raw_results, filter, p.adj_bonferroni < 0.1) %>% 
  lapply(mutate_if, is.numeric, round, 4) %>% 
  lapply(reactable, sortable = TRUE, filterable = TRUE)
```

```{r}
tmp$DoD
```

```{r}
tmp$NGRC
```

```{r}
tmp$UAB
```

### Conclusions of approach 1 (ClusterSum)

  * Within dataset:
    - There were no Bonferroni-adjusted interaction associations
  * Across datasets:
    - No overlap of the few associations marginally significant considering raw
    p-values.
    
## Approach 2 - Regression of each microbe, followed by Microbe Set Enrichment analysis

### LinDA association only

```{r}
approach2_interaction_linda <- list(
  DoD = read_tsv("output_DoD/Q3/02_LinDA_regression_dosage_in_Controls.tsv") %>% mutate(Dataset = "DoD"),
  NGRC = read_tsv("output_NGRC/Q3/02_LinDA_regression_dosage_in_Controls.tsv") %>% mutate(Dataset = "NGRC"),
  UAB = read_tsv("output_UAB/Q3/02_LinDA_regression_dosage_in_Controls.tsv") %>% mutate(Dataset = "UAB")
) %>% 
  lapply(relocate, Dataset) %>% 
  lapply(filter, grepl("interactionPD", covariate)) 

```

```{r, results="asis"}

tables_as_list <- approach2_interaction_linda %>% 
  lapply(filter, padj < 0.1) %>% 
  lapply(mutate_if, is.numeric, round, 4) %>% 
  lapply(select, -lfcSE, -reject, -df, -covariate) %>% 
  lapply(reactable, sortable = TRUE, filterable = TRUE)

```

```{r}
tables_as_list$DoD
```

```{r}
tables_as_list$NGRC
```

```{r}
tables_as_list$UAB
```

Are there cases where the same variant:PD interaction term is significantly 
associated with the same bacterium in more than one dataset?

```{r}
tmp <- approach2_interaction_linda %>% 
  lapply(filter, padj < 0.1) %>% 
  reduce(full_join, by = c("taxon", "variantName"))

associations_replic <- tmp[rowSums(is.na(select(tmp, contains("log2FoldChange")))) < 2,]

linda_intersection_also_nonsignif <- approach2_interaction_linda %>% 
  lapply(filter, variantName %in% associations_replic$variantName, taxon %in% associations_replic$taxon) %>% 
  reduce(rbind.data.frame) %>% 
  mutate(alpha = ifelse(padj < 0.1, 1, 0.5))

reactable(linda_intersection_also_nonsignif)
```

There are not replications of associations across datasets, but multiple ones 
in each separately.

### Microbe Set Enrichment Analysis with LinDA `statistic`s

```{r}
approach2_linda_MSEA <- list(
  DoD = read_tsv("output_DoD/Q3/03_full_MSEA_results_variant_in_Controls.tsv") %>% mutate(Dataset = "DoD"),
  NGRC = read_tsv("output_NGRC/Q3/03_full_MSEA_results_variant_in_Controls.tsv") %>% mutate(Dataset = "NGRC"),
  UAB = read_tsv("output_UAB/Q3/03_full_MSEA_results_variant_in_Controls.tsv") %>% mutate(Dataset = "UAB")
) %>% 
  lapply(relocate, Dataset)
```

Below is the table of union of Microbe sets enriched or depleted across
the three datasets.

```{r}
approach2_linda_MSEA  %>%  
  lapply(filter, padj < 0.1) %>% 
  lapply(select, -leadingEdge) %>% 
  reduce(full_join, by = c("clusterName", "snp_variant")) %>% 
  mutate_if(is.numeric, round, 4) %>% 
  reactable()
```

To better view the degree of replication, let's plot all and show where
FDR q-values are less than 0.1.

```{r, fig.caption = "Plot of variants with at least 2 datasets associated in MSEA using linda t-statistics on the interaction term. Semi-transparent bars indicate datasets with non-significant enrichment or depletion for that specific bacterial cluster and variant. Dotted lines indicate an arbitrary threshold of microbe set enrichment at |NES| > 1.2"}
tmp <- approach2_linda_MSEA %>% 
  lapply(filter, padj < 0.1) %>% 
  reduce(full_join, by = c("clusterName", "snp_variant"))

# restrict to rows that have at least 2 datasets replicated
# that is the row must contain between 0 and 1 NAs
associations_replic <- tmp[rowSums(is.na(select(tmp, contains("NES")))) < 2,]


MSEA_intersection_also_nonsignif <- approach2_linda_MSEA %>% 
  lapply(filter, snp_variant %in% associations_replic$snp_variant, clusterName %in% associations_replic$clusterName) %>% 
  reduce(rbind.data.frame) %>% 
  mutate(alpha = ifelse(padj < 0.1, 1, 0.5))
  
NES_barplot <- ggplot(MSEA_intersection_also_nonsignif, aes(x = clusterName, y = NES, fill = Dataset)) + 
  geom_col(position = position_dodge(), mapping = aes(alpha = alpha)) +
facet_wrap(~snp_variant) + 
  theme_light() + 
  geom_hline(yintercept = 1.2, col = "black", lty = "dotted") +
  geom_hline(yintercept = -1.2, col = "black", lty = "dotted") + 
  guides(alpha = "none") + 
  scale_fill_manual(values = ggsci::pal_jco()(4)[c(4,1,2)])

NES_barplot
```

