---
title: "WallenZD_2022 data preparation"
author: "Giacomo Antonello, Zachary D. Wallen, Chad Murchinson, Levi Waldron, Haydeh Payami"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: show
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
```

```{r, include=FALSE}
# Check and install necessary CRAN packages
CRAN_required_packages <- c("tidyverse", "gwasrapidd", "DT", "kableExtra", "data.table", "ggpubr", "BiocManager", "readxl","kableExtra")

install.packages(
  CRAN_required_packages[!(CRAN_required_packages %in% installed.packages())] 
)

# Check and install necessary Bioconductor packages
BioC_required_packages <- c("Maaslin2", "ANCOMBC", "biomaRt", "mia", "phyloseq", "microbiome","SIAMCAT")
BioC_required_packages <- BioC_required_packages[!(BioC_required_packages %in% installed.packages())]

if(length(BioC_required_packages) > 0) {
  BiocManager::install(BioC_required_packages[!(BioC_required_packages %in% installed.packages())],
                       ask = FALSE,
                       update = TRUE)
}


# load basic data manipulation packages
library(phyloseq)
library(DT)
library(kableExtra)
library(MultiAssayExperiment)
library(tidyverse)
# load previously wrangled data
load("../Data/wallenZD_data.mae.RData")

rm(BioC_required_packages,CRAN_required_packages)
gc(verbose = FALSE)

sapply(list.files("~/bin/biobakeryUtils/", full.names = TRUE), source)
```

# Directory and data preparation

First thing is clone the repository I made for the Payami Lab and add make sure 
that the `.gitignore` file has exceptions for unwanted data sharing into the 
repository. `NB: This has to be done manually`

```{bash, eval=FALSE}
cd ~

git clone git@github.com:ASAP-MAC/PayamiH.git

cd PayamiH
```

# Work directories and genetic data

1 - Data directories generation. 
**Maker sure you start from the base Git directory**

```{bash setup data directory, eval=FALSE}
# TOPMed data
mkdir -p Data/TOPMed_imputed_genotypes
cp -r /data/project/payami-lab/payamilab/TOPMed_imputed_genotypes_2022/* Data/TOPMed_imputed_genotypes

# Microbiome data
mkdir -p Data/Zenodo
wget https://zenodo.org/records/7246185/files/Source_Data_24Oct2022.xlsx?download=1 -O /Data/Zenodo/Source_Data_24Oct2022.xlsx

```

2 - bcftools setup

```{bash setup micromamba, eval=FALSE}
cd ~

curl -Ls https://micro.mamba.pm/api/micromamba/linux-64/latest | tar -xvj micromamba/bin
./bin/micromamba shell init -s bash -r ~/micromamba

micromamba config append channels bioconda

micromamba create -n geneticTools -c bioconda

micromamba activate geneticTools

micromamba install bcftools

```

3 - Indexing of VCF file for fast querying (necessary for bcftools)

```{bash tabix vcf files, eval=FALSE}
cd ~/PayamiH/Data/TOPMed_imputed_genotypes/

for chr in `ls *.dose.vcf.gz`; do sbatch -c1 --mem 8G -pexpress --wrap "bcftools index -c ${chr}"; done
```

4 - Extract samples list to query with `bcftools query`, this is done considering the following:

  1. The only samples needed for the analysis are those beginning with `D`
  2. Keep only samples from point 1 that present across all VCF files. This is 
  redundant but done anyway.

```{bash, eval=FALSE}
cd ~/PayamiH/Data

mkdir sampleNames/

find TOPMed_imputed_genotypes_2022/ -type f -name "*.vcf.gz" -exec bcftools query -l {} \; >> sampleNames/sampleNamesAll.txt

cat sampleNames/sampleNamesAll.txt | egrep "^D" | sort | uniq -c | sed -e "s/^[\ ]*//" | tr " " "\t" | awk '{if ($1 == 23) print $2}' > sampleNames/sampleNamesPD.txt

cat sampleNames/sampleNamesPD.txt | tr "\n" "," | sed "s/\,$//g" > sampleNames/sampleNamesPD_oneLine.txt
```

The last step is important for bcftools to work properly in querying samples. 

# Project setup, PD genetics and gut microbiome

```{bash, eval=FALSE}
cd ~/PayamiH

mkdir -p project_1_MicroGenPD/output/

```

# Download all unique GWAS catalog variants with PD (based on `rsID`)

Now we should download the genetic variants that we are interested in, being
those associated with `PD`. Notice that there are duplicated `rsID`s depending
on how many times they have been entried, hence reported. We are interested in 
having them only once.

```{r, eval=!interactive()}
# retrieve associations with GWAS catalog traits
parkinson_gwasCatalog.gwasrapidd <- gwasrapidd::get_associations(efo_id = "MONDO_0005180")

parkinson_variants_gwas_catalog.df <- full_join(
  select(parkinson_gwasCatalog.gwasrapidd@associations, 
         association_id, or_per_copy_number, beta_unit),
  select(parkinson_gwasCatalog.gwasrapidd@risk_alleles, 
         association_id, variant_id, risk_allele), 
  by = "association_id") %>% 
  # remove duplicated lines, preferring those where a rik allele has been reported
  arrange(risk_allele) %>%       
  distinct(variant_id, .keep_all = TRUE)

ensembl <- biomaRt::useEnsembl(biomart = "snps", dataset = "hsapiens_snp")

rsid_to_CHR_POS <- biomaRt::getBM(
  attributes = c("refsnp_id", "chr_name", "chrom_start", "chrom_end", "allele", "associated_gene"), 
      filters = "snp_filter", 
      values = parkinson_variants_gwas_catalog.df$variant_id, 
      mart = ensembl) %>% 
  rename(
    variant_id = refsnp_id
  ) %>%
  #keep only well defined chromosome scaffolds, including sex chromosome
  filter(
    !startsWith(chr_name, "H")
  ) %>% 
  mutate(
    chr_name = factor(paste0("chr", chr_name), levels = as.character(paste0("chr", c(1:22, "X", "Y"))))
  ) %>% 
  select(
    CHROM = chr_name, 
    CHROM_START = chrom_start, 
    CHROM_END = chrom_end, 
    ALLELE = allele, 
    RSID = variant_id,
    GENE = associated_gene
    ) %>% 
  # same trick, to be sure we have only variants with gene names associated, wherever possible
  arrange(rowSums(is.na(.))) %>%       
  distinct(RSID, .keep_all = TRUE)

write_tsv(rsid_to_CHR_POS, file = "GWAS_catalog_PD_variants/PD_variants_GWAS_catalog.tsv")

for(chr in unique(rsid_to_CHR_POS$CHROM)){
  writeLines(text = paste0(filter(rsid_to_CHR_POS, CHROM == chr)$CHROM, ":", filter(rsid_to_CHR_POS, CHROM == chr)$CHROM_START,"-", filter(rsid_to_CHR_POS, CHROM == chr)$CHROM_END), con = paste0("GWAS_catalog_PD_variants/", chr, "_variants_Tabix.txt"))
}

datatable(rsid_to_CHR_POS, caption = "Variants available in the GWAS catalog made unique based on the rsID. Additionally, Gene annotations were given preference over no annotation. Hence, they should be treated with caution. All annotations were done using the ensembl annotation resources in `biomaRt`.")
```

```{bash, eval=FALSE}
# from the PayamiH as base directory, do the following
cd Data/GWAS_catalog_PD_variants 

for chr in $(ls | grep -oE "^(chr[0-9]*|chrX)"); do bcftools query -s $(cat ../sampleNames/sampleNamesPD_oneLine.txt) -H -f "%CHROM\t%POS\t%REF\t%ALT\t%INFO[\t%GT]\n" -r $(tr "\n" "," < ${chr}_variants_Tabix.txt | sed "s/,$//") ../TOPMed_imputed_genotypes/${chr}.dose.vcf.gz | sed -e "s/# //g" -e "s/:GT//g" -e "s/\[[0-9]*\]//g" -e "s/(null)/INFO/g" > queried_variants_PD/${chr}_pd_Genotypes.txt; done

for chr in $(ls | grep -oE "^(chr[0-9]*|chrX)"); do bcftools query -s $(cat ../sampleNames/sampleNamesPD_oneLine.txt) -H -f "%CHROM\t%POS\t%REF\t%ALT\t%INFO[\t%DS]\n" -r $(tr "\n" "," < ${chr}_variants_Tabix.txt | sed "s/,$//") ../TOPMed_imputed_genotypes/${chr}.dose.vcf.gz | sed -e "s/# //g" -e "s/:DS//g" -e "s/\[[0-9]*\]//g" -e "s/(null)/INFO/g" > queried_variants_PD/${chr}_pd_Dosages.txt; done

```

# Retrieve PD associated variants in SNCA and HLA genes in Wallen genetic data

More specifically, we will work with dosages.

Additionally, they must be available in WallenZD.

```{r load variants dosages and hardcalls}
gwasCatalog_variants_dosages_All <- data.table::fread("GWAS_catalog_PD_variants/PD_variants_GWAS_catalog.tsv", data.table = FALSE) %>% 
  select(CHROM, POS = CHROM_START, GENE, RSID) %>% 
  # force unique chrom and pos and group Genes and rsid together, this is redundant
  group_by(CHROM, POS) %>% 
  reframe(
    GENE = paste(GENE, collapse = ";"),
    RSID = paste(RSID, collapse = ";")
  ) %>% as.data.frame()

gwasCatalog_variants_dosages_in_WallenZD.df <- list.files(path = "GWAS_catalog_PD_variants/queried_variants_PD/", pattern = "*Dosages.txt", full.names = T) %>% 
  sapply(data.table::fread, USE.NAMES = TRUE, simplify = FALSE) %>% 
  lapply(inner_join, gwasCatalog_variants_dosages_All, by = c("CHROM", "POS")) %>% 
  lapply(relocate, RSID, GENE, .after = ALT) %>% 
  lapply(function(x) set_names(x, str_remove(colnames(x), "_.*"))) %>% 
  purrr::reduce(rbind) %>% 
  as.data.frame()

gwasCatalog_variants_hardcalls_in_WallenZD.df <- list.files(path = "GWAS_catalog_PD_variants/queried_variants_PD/", pattern = "*Genotypes.txt", full.names = T) %>% 
  sapply(data.table::fread, USE.NAMES = TRUE, simplify = FALSE) %>% 
  lapply(inner_join, gwasCatalog_variants_dosages_All, by = c("CHROM", "POS")) %>% 
  lapply(relocate, RSID, GENE, .after = ALT) %>% 
  lapply(function(x) set_names(x, str_remove(colnames(x), "_.*"))) %>% 
  purrr::reduce(rbind) %>% 
  as.data.frame()

# summary table of variants found in the dataset
nVars_by_genes_All <- gwasCatalog_variants_dosages_All %>% 
  mutate(
    `SNCA or HLA` = grepl("SNCA|HLA", GENE), 
    `LRRK2 or GBA` = grepl("LRRK2|GBA", GENE), 
    OTHER = !grepl("SNCA|HLA|GBA|LRRK2", GENE)) %>% 
  .[,(ncol(.)-2):ncol(.)] %>% 
  colSums()

nVars_by_genes_Wallen <- gwasCatalog_variants_dosages_in_WallenZD.df %>% 
  mutate(
    `SNCA or HLA` = grepl("SNCA|HLA", GENE), 
    `LRRK2 or GBA` = grepl("LRRK2|GBA", GENE), 
    OTHER = !grepl("SNCA|HLA|GBA|LRRK2", GENE)) %>% 
  .[,(ncol(.)-2):ncol(.)] %>% 
  colSums()

cbind(c("Tot" = sum(nVars_by_genes_All), nVars_by_genes_All), 
      c("Tot" = sum(nVars_by_genes_Wallen), nVars_by_genes_Wallen)) %>% 
  
  magrittr::set_colnames(c("GWAS Catalog", "WallenZD")) %>% 
  kableExtra::kbl(caption = "Comparison of genetic variants associated with PD found in the GWAS catalog and in Wallen imputed genotypes") %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::add_header_above(c(" " = 1, "N. unique variants associated with PD" = 2))

```


At this step, we have generated a set of genotypes to test against the microbiome
as found in the GWAS catalog and our genotypes. It is a `samples x genotypes` matrix.

# WallenZD data preparation into `MultiAssayExperiment`

```{r}

theme_set(theme_light())
# load and wrangle available data

# 1 - metaphlan relative abundances and estimated counts
WallenZD_2022_metaphlan_relab <- readxl::read_xlsx("Zenodo/Source_Data_24Oct2022.xlsx", sheet = "metaphlan_rel_ab")
WallenZD_2022_metaphlan_counts <- readxl::read_xlsx("Zenodo/Source_Data_24Oct2022.xlsx", sheet = "metaphlan_counts")
# 1.1 - clean metaphlan tables
WallenZD_2022_metaphlan_relab_wrangled <- wrangle_metaphlan(WallenZD_2022_metaphlan_relab, "Species")
WallenZD_2022_metaphlan_counts_wrangled <- wrangle_metaphlan(WallenZD_2022_metaphlan_counts, "Species")
WallenZD_2022_metaphlan_counts_wrangled$profiles <- round(WallenZD_2022_metaphlan_counts_wrangled$profiles, 0)

# 2 - HUMaNN profiles
# 2.1 KO tables
WallenZD_2022_humann_KO <- readxl::read_xlsx("Zenodo/Source_Data_24Oct2022.xlsx", sheet = "humann_KO_group_counts") %>% 
  wrangle_humann(input_type = "ko")
WallenZD_2022_humann_KO$profiles <- round(WallenZD_2022_humann_KO$profiles, 0)
WallenZD_2022_humann_pathways <- readxl::read_xlsx("Zenodo/Source_Data_24Oct2022.xlsx", sheet = "humann_pathway_counts") %>% wrangle_humann(input_type = "pathways")

# 2.2 - Pathways
WallenZD_2022_humann_pathways$profiles <- round(WallenZD_2022_humann_pathways$profiles, 0)

# 3 - sample-participant metadata
WallenZD_2022_metadata <- readxl::read_xlsx("Zenodo/Source_Data_24Oct2022.xlsx", sheet = "subject_metadata") %>% 
  as.data.frame()
rownames(WallenZD_2022_metadata) <- WallenZD_2022_metadata$sample_name

# reorder colnames so they all match the metadata
samples_intersect <- purrr::reduce(list(WallenZD_2022_metaphlan_relab_wrangled$profiles,
WallenZD_2022_metaphlan_counts_wrangled$profiles,
WallenZD_2022_humann_KO$profiles,
WallenZD_2022_humann_pathways$profiles) %>% lapply(colnames), intersect) 

WallenZD_2022_metaphlan_relab_wrangled$profiles <- WallenZD_2022_metaphlan_relab_wrangled$profiles[, samples_intersect]
WallenZD_2022_metaphlan_counts_wrangled$profiles <- WallenZD_2022_metaphlan_counts_wrangled$profiles[, samples_intersect]
WallenZD_2022_humann_KO$profiles <- WallenZD_2022_humann_KO$profiles[, samples_intersect]
WallenZD_2022_humann_pathways$profiles <- WallenZD_2022_humann_pathways$profiles[, samples_intersect]

sample_metadata.df <- WallenZD_2022_metadata[samples_intersect,] %>% 
  mutate(seqs_scaled = scale(total_sequences))

```

```{r}
# manipulate genetics data a bit, to make it more usable in the SummarizedExperiment
# context

tmp <- gwasCatalog_variants_dosages_in_WallenZD.df
samples_intersect <- intersect(colnames(tmp), rownames(sample_metadata.df))
metadata_cols_genetics <- grep("^D", colnames(tmp), value = TRUE, invert = T)
variants_made_unique <- make.unique(tmp$RSID)
rownames(tmp) <- variants_made_unique

genetics_matrix_dosages <- as.matrix(tmp[variants_made_unique, samples_intersect])

tmp <- gwasCatalog_variants_hardcalls_in_WallenZD.df
samples_intersect <- intersect(colnames(tmp), rownames(sample_metadata.df))
metadata_cols_genetics <- grep("^D", colnames(tmp), value = TRUE, invert = T)
variants_made_unique <- make.unique(tmp$RSID)
rownames(tmp) <- variants_made_unique

genetics_matrix_hardcalls <- as.matrix(tmp[variants_made_unique, samples_intersect])


genetics_metadata <- tmp[variants_made_unique, metadata_cols_genetics]
sample_metadata_genetics <- sample_metadata.df[samples_intersect,]

library(MultiAssayExperiment, quietly = TRUE, verbose = FALSE)

wallenZD_summarizedExperiments.list <- list(
  genotypes_dosages.se = SummarizedExperiment(
    assays = list("genotypes_dosages" = genetics_matrix_dosages, "genotypes_hardcalls" = genetics_matrix_hardcalls),
    rowData = genetics_metadata,
  colData = sample_metadata_genetics
  ),
  
  genotypes_hardcalls.se = SummarizedExperiment(
    assays = list("genotypes_hardcalls" = genetics_matrix_hardcalls),
    rowData = genetics_metadata,
  colData = sample_metadata_genetics
  ), 
  
  metaphlan_RelAbund.se = SummarizedExperiment(
    assays = list("metaphlan_RelAbund" = WallenZD_2022_metaphlan_relab_wrangled$profiles),
    rowData =  WallenZD_2022_metaphlan_relab_wrangled$taxonomies,
    colData = sample_metadata.df
  ),
  
  metaphlan_Counts.se = SummarizedExperiment(
    assays = list("metaphlan_Counts" = WallenZD_2022_metaphlan_counts_wrangled$profiles),
    rowData =  WallenZD_2022_metaphlan_counts_wrangled$taxonomies,
    colData = sample_metadata.df
  ),
  
  humann_ko.se = SummarizedExperiment(
    assays = list("humann_ko" = WallenZD_2022_humann_KO$profiles),
    rowData =  WallenZD_2022_humann_KO$description,
    colData = sample_metadata.df
  ),
  
  humann_pathways.se = SummarizedExperiment(
    assays = list("humann_pathways" = WallenZD_2022_humann_pathways$profiles),
    rowData =  WallenZD_2022_humann_pathways$description,
    colData = sample_metadata.df
  )
)

assays_vec <- lapply(wallenZD_summarizedExperiments.list, colnames) %>% 
  lapply(length)

assays_vec <- lapply(names(assays_vec), function(x) rep(x, assays_vec[[x]])) %>% 
  purrr::reduce(c) %>% 
  factor(levels = names(assays_vec))

assays_sample_map <- data.frame(
  assay = assays_vec,
  primary = lapply(wallenZD_summarizedExperiments.list, colnames) %>% purrr::reduce(c),
  colname = lapply(wallenZD_summarizedExperiments.list, colnames) %>% purrr::reduce(c)
)

wallenZD_data.mae <- MultiAssayExperiment(
  experiments = wallenZD_summarizedExperiments.list,
  sampleMap = assays_sample_map,
  colData = sample_metadata.df
)
```

```{r clean environment and memory for next analyses}

save(wallenZD_data.mae, file = "wallenZD_data.mae.RData")

```

```{r}
wallenZD_data.mae
```

```{r}
assays(wallenZD_data.mae)
```

