---
title: "NGRC data preparation into MultiAssayExperiment format"
author: "Giacomo Antonello, Zachary Wallen, Charles Murchinson, Levi Waldron, Haydeh Payami"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: show
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
---
Latest render: `r format(Sys.time(), "Date: %Y-%b-%d Time: %H:%M:%S")`

```{r, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
```

```{r, include=FALSE}
# Check and install necessary CRAN packages
CRAN_required_packages <- c("tidyverse", "gwasrapidd", "DT", "kableExtra", "data.table", "ggpubr", "BiocManager", "readxl","kableExtra")

install.packages(
  CRAN_required_packages[!(CRAN_required_packages %in% installed.packages())] 
)

# Check and install necessary Bioconductor packages
BioC_required_packages <- c("Maaslin2", "ANCOMBC", "biomaRt", "mia", "phyloseq", "microbiome","SIAMCAT")
BioC_required_packages <- BioC_required_packages[!(BioC_required_packages %in% installed.packages())]

if(length(BioC_required_packages) > 0) {
  BiocManager::install(BioC_required_packages[!(BioC_required_packages %in% installed.packages())],
                       ask = FALSE,
                       update = TRUE)
}


# load basic data manipulation packages
library(phyloseq)
library(DT)
library(kableExtra)
library(MultiAssayExperiment)
library(tidyverse)

rm(BioC_required_packages,CRAN_required_packages)
gc(verbose = FALSE)

for(f in list.files("~/bin/biobakeryUtils/", full.names = TRUE)){
  source(f)
}
```

# Prepare microbiome-related data

```{r}
# get sample metadata
NGRC_metadata <- as.data.frame(readxl::read_xlsx("/data/project/payami-lab/payamilab/datasets_sample_renaming/NGRC_new_SourceFile.xlsx", sheet = "subject_metadata"))
rownames(NGRC_metadata) <- NGRC_metadata$SubjectID

# get Metaphlan relative abundances
NGRC_metaphlan_relab <- readxl::read_xlsx("/data/project/payami-lab/payamilab/datasets_sample_renaming/NGRC_new_SourceFile.xlsx", sheet = "metaphlan_rel_ab") %>% 
  wrangle_metaphlan("Species")

# get Metaphlan estimated counts
NGRC_metaphlan_counts <- readxl::read_xlsx("/data/project/payami-lab/payamilab/datasets_sample_renaming/NGRC_new_SourceFile.xlsx", sheet = "metaphlan_counts") %>% wrangle_metaphlan("Species")
NGRC_metaphlan_counts$profiles <- round(NGRC_metaphlan_counts$profiles, 0)

# get HUMAnN KO tables
NGRC_humann_KO <- readxl::read_xlsx("/data/project/payami-lab/payamilab/datasets_sample_renaming/NGRC_new_SourceFile.xlsx", sheet = "humann_KO_group_counts") %>% 
  wrangle_humann(input_type = "ko")
NGRC_humann_KO$profiles <- round(NGRC_humann_KO$profiles, 0)

# get HUMAnN Pathway tables
NGRC_humann_pathways <- readxl::read_xlsx("/data/project/payami-lab/payamilab/datasets_sample_renaming/NGRC_new_SourceFile.xlsx", sheet = "humann_pathway_counts") %>% wrangle_humann(input_type = "pathways")
NGRC_humann_pathways$profiles <- round(NGRC_humann_pathways$profiles, 0)

# find samples intersection across experiments and metadata

samples_per_data_piece <- list(
  "Metaphlan (Relative Abundance)" = colnames(NGRC_metaphlan_relab$profiles),
  "Metaphlan (Counts)" = colnames(NGRC_metaphlan_counts$profiles),
  "HUMAnN3 (Kegg Orthology)" = colnames(NGRC_humann_KO$profiles),
  "HUMAnN3 (MetaCyc Pathways)" = colnames(NGRC_humann_pathways$profiles),
  "Metadata" = rownames(NGRC_metadata)
  )

samples_intersect <- reduce(samples_per_data_piece, intersect)

# load samples to exclude (not necessary in DoD, but necessary in UAB and NGRC)
samples_to_exclude.df <- readxl::read_xlsx("/data/project/payami-lab/payamilab/0_Source_Data/mapping_and_exclusion_lists/exclusions NGRC UAB_ZW_HP.xlsx", sheet = "R-friendly list")

#samples_intersect_after_exclusion <- samples_intersect[!(samples_intersect %in% samples_to_exclude.df$sample_name)]
samples_intersect_after_exclusion <- samples_intersect

# reorder colnames so they all match the metadata

NGRC_metaphlan_relab$profiles <- NGRC_metaphlan_relab$profiles[, samples_intersect_after_exclusion]
NGRC_metaphlan_counts$profiles <- NGRC_metaphlan_counts$profiles[, samples_intersect_after_exclusion]
NGRC_humann_KO$profiles <- NGRC_humann_KO$profiles[, samples_intersect_after_exclusion]
NGRC_humann_pathways$profiles <- NGRC_humann_pathways$profiles[, samples_intersect_after_exclusion]

sample_metadata.df <- NGRC_metadata[samples_intersect_after_exclusion,] %>% 
  mutate(seqs_scaled = scale(total_sequences))

samples_to_exclude.df %>% 
  filter(sample_name %in% samples_intersect) %>% 
  kbl(caption = "List of samples excluded from the NGRC dataset.")
```

## Prepare genotype data

Both dosages and hardcalls.

```{r load variants dosages and hardcalls}
# Load all variants found in GWAS catalog under our criteria

gwasCatalog_variants_dosages_All <- data.table::fread("GWAS_catalog_PD_variants/PD_variants_GWAS_catalog.tsv", data.table = FALSE) %>% 
  rename(POS = CHROM_START,
         PD.RISK.ALLELE = PD_RISK_ALLELE) %>% 
  select(-CHROM_END) %>% 
  as.data.frame()

# Load variants found in TOPMed datasets, with all samples available (> 5k)

gwasCatalog_variants_dosages_in_TOPMed.df <- list.files(path = "GWAS_catalog_PD_variants/queried_variants_PD/", pattern = "*Dosages.txt", full.names = T) %>% 
  sapply(data.table::fread, USE.NAMES = TRUE, simplify = FALSE) %>% 
  lapply(inner_join, gwasCatalog_variants_dosages_All, by = c("CHROM", "POS")) %>% 
  lapply(relocate, PD.RISK.ALLELE, RSID, GENE, .after = ALT) %>% 
  purrr::reduce(rbind) %>% 
  # wrangle INFO column for further info
  separate_rows(INFO, sep = ";") %>%
  separate(INFO, into = c("key", "value"), sep = "=", fill = "right") %>%
  pivot_wider(names_from = key, values_from = value) %>%
  mutate(across(everything(), ~replace_na(.x, NA)),
         IMPUTED = ifelse(
           is.na(ER2), "TRUE", "FALSE"), 
         TYPED = NULL) %>% 
  relocate(
    INFO_AF = AF, INFO_MAF = MAF, INFO_R2 = R2, INFO_ER2 = ER2, INFO_IMPUTED = IMPUTED, 
    .after = GENE
  ) %>%
  mutate(across(contains("INFO"), as.numeric)) %>% 
  # as last step to be able to assign row names 2 chunks later
  as.data.frame()
  
# load hardcalls 
  
gwasCatalog_variants_hardcalls_in_TOPMed.df <- list.files(path = "GWAS_catalog_PD_variants/queried_variants_PD/", pattern = "*Genotypes.txt", full.names = T) %>% 
  sapply(data.table::fread, USE.NAMES = TRUE, simplify = FALSE) %>% 
  lapply(inner_join, gwasCatalog_variants_dosages_All, by = c("CHROM", "POS")) %>% 
  lapply(relocate, PD.RISK.ALLELE, RSID, GENE, .after = ALT) %>% 
  purrr::reduce(rbind) %>% 
  # wrangle INFO column for further info
  separate_rows(INFO, sep = ";") %>%
  separate(INFO, into = c("key", "value"), sep = "=", fill = "right") %>%
  pivot_wider(names_from = key, values_from = value) %>%
  mutate(across(everything(), ~replace_na(.x, NA)),
         IMPUTED = ifelse(
           is.na(ER2), "TRUE", "FALSE"), 
         TYPED = NULL) %>% 
  
  relocate(
    INFO_AF = AF, INFO_MAF = MAF, INFO_R2 = R2, INFO_ER2 = ER2, INFO_IMPUTED = IMPUTED, 
    .after = GENE
  ) %>%
  mutate(across(contains("INFO"), as.numeric)) %>% 
  # as last step to be able to assign row names 2 chunks later
  as.data.frame()

# summary table of variants found in the dataset
nVars_by_genes_All <- gwasCatalog_variants_dosages_All %>% 
  mutate(
    `SNCA or HLA` = grepl("SNCA|HLA", GENE), 
    `LRRK2 or GBA` = grepl("LRRK2|GBA", GENE), 
    OTHER = !grepl("SNCA|HLA|GBA|LRRK2", GENE)) %>% 
  .[,(ncol(.)-2):ncol(.)] %>% 
  colSums()

nVars_by_genes_TOPMed <- gwasCatalog_variants_dosages_in_TOPMed.df %>% 
  mutate(
    `SNCA or HLA` = grepl("SNCA|HLA", GENE), 
    `LRRK2 or GBA` = grepl("LRRK2|GBA", GENE), 
    OTHER = !grepl("SNCA|HLA|GBA|LRRK2", GENE)) %>% 
  .[,(ncol(.)-2):ncol(.)] %>% 
  colSums()

cbind(c("Tot" = sum(nVars_by_genes_All), nVars_by_genes_All), 
      c("Tot" = sum(nVars_by_genes_TOPMed), nVars_by_genes_TOPMed)) %>% 
  
  magrittr::set_colnames(c("GWAS Catalog", "TOPMed imputed PayamiLab samples")) %>% 
  kableExtra::kbl(caption = "Comparison of genetic variants associated with PD found in the GWAS catalog and in TOPMed imputed genotypes") %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::add_header_above(c(" " = 1, "N. unique variants associated with PD" = 2))

```

At this step, we have generated a set of genotypes to test against the microbiome
as found in the GWAS catalog and our genotypes. It is a `samples x genotypes` matrix.

```{r}
# manipulate genetics data a bit, to make it more usable in the SummarizedExperiment
# context
variant_metadata_cols <- colnames(gwasCatalog_variants_dosages_in_TOPMed.df)[1:max(which(grepl("INFO", colnames(gwasCatalog_variants_dosages_in_TOPMed.df))))]

tmp <- gwasCatalog_variants_dosages_in_TOPMed.df
samples_intersect_genetics <- intersect(colnames(tmp), rownames(sample_metadata.df))
metadata_cols_genetics <- colnames(tmp)[1:max(grep("INFO_", colnames(tmp)))]
variants_made_unique <- make.unique(tmp$RSID)
rownames(tmp) <- variants_made_unique

genetics_matrix_dosages <- as.matrix(tmp[variants_made_unique, samples_intersect_genetics])

tmp <- gwasCatalog_variants_hardcalls_in_TOPMed.df
rownames(tmp) <- variants_made_unique

genetics_matrix_hardcalls <- as.matrix(tmp[variants_made_unique, samples_intersect_genetics])


genetics_metadata <- tmp[variants_made_unique, metadata_cols_genetics]
sample_metadata_genetics <- sample_metadata.df[samples_intersect_genetics,]
```

## Generate a `MultiAssayExperiment` object with all data together

```{r}

library(MultiAssayExperiment, quietly = TRUE, verbose = FALSE)

summarizedExperiments.list <- list(
  genotypes_dosages.se = SummarizedExperiment(
    assays = list("genotypes_dosages" = genetics_matrix_dosages, "genotypes_hardcalls" = genetics_matrix_hardcalls),
    rowData = genetics_metadata,
  colData = sample_metadata_genetics
  ),
  
  genotypes_hardcalls.se = SummarizedExperiment(
    assays = list("genotypes_hardcalls" = genetics_matrix_hardcalls),
    rowData = genetics_metadata,
  colData = sample_metadata_genetics
  ), 
  
  metaphlan_RelAbund.se = SummarizedExperiment(
    assays = list("metaphlan_RelAbund" = NGRC_metaphlan_relab$profiles),
    rowData =  NGRC_metaphlan_relab$taxonomies,
    colData = sample_metadata.df
  ),
  
  metaphlan_Counts.se = SummarizedExperiment(
    assays = list("metaphlan_Counts" = NGRC_metaphlan_counts$profiles),
    rowData =  NGRC_metaphlan_counts$taxonomies,
    colData = sample_metadata.df
  ),
  
  humann_KO.se = SummarizedExperiment(
    assays = list("humann_KO" = NGRC_humann_KO$profiles),
    rowData =  NGRC_humann_KO$description,
    colData = sample_metadata.df
  ),
  
  humann_pathways.se = SummarizedExperiment(
    assays = list("humann_pathways" = NGRC_humann_pathways$profiles),
    rowData =  NGRC_humann_pathways$description,
    colData = sample_metadata.df
  )
)

assays_vec <- lapply(summarizedExperiments.list, colnames) %>% 
  lapply(length)

assays_vec <- lapply(names(assays_vec), function(x) rep(x, assays_vec[[x]])) %>% 
  purrr::reduce(c) %>% 
  factor(levels = names(assays_vec))

assays_sample_map <- data.frame(
  assay = assays_vec,
  primary = lapply(summarizedExperiments.list, colnames) %>% purrr::reduce(c),
  colname = lapply(summarizedExperiments.list, colnames) %>% purrr::reduce(c)
)

NGRC_data.mae <- MultiAssayExperiment(
  experiments = summarizedExperiments.list,
  sampleMap = assays_sample_map,
  colData = sample_metadata.df
)
```

```{r clean environment and memory for next analyses}

save(NGRC_data.mae, file = "NGRC_data.mae.RData")

```

# Final data object generated - a quick look

```{r}
NGRC_data.mae
```

# Samples not fully retrieved

## Samples excluded at each step

```{r}

cbind(
  sapply(samples_per_data_piece, length),
  rep(length(samples_intersect), length(samples_per_data_piece)),
  rep(length(samples_intersect_after_exclusion), length(samples_per_data_piece)),
  rep(length(samples_intersect_genetics), length(samples_per_data_piece))
  ) %>%
  magrittr::set_colnames(c("Unifiltered", "With Microbiome", "After sample exclusion", "Also with Genetic data")) %>% 
  kbl(caption = "Number of samples in raw input data (Unfiltered) and after finding the samples in common (Intersection)")

```


## Microbiome samples not retrieved in Genetic data

```{r}
colnames(NGRC_data.mae[["metaphlan_RelAbund.se"]])[!(colnames(NGRC_data.mae[["metaphlan_RelAbund.se"]]) %in% colnames(NGRC_data.mae[["genotypes_dosages.se"]]))]
```


# Session Info

```{r}
sessionInfo()
```

