---
title: "WallenZD_2022 data preparation"
author: "Giacomo Antonello, Zachary D. Wallen, Chad Murchinson, Levi Waldron, Haydeh Payami"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: show
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
---
Latest render: `r format(Sys.time(), "Date: %Y-%b-%d Time: %H:%M:%S")`

```{r, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
```

```{r, include=FALSE}
# Check and install necessary CRAN packages
CRAN_required_packages <- c("tidyverse", "gwasrapidd", "DT", "kableExtra", "data.table", "ggpubr", "BiocManager", "readxl","kableExtra")

install.packages(
  CRAN_required_packages[!(CRAN_required_packages %in% installed.packages())] 
)

# Check and install necessary Bioconductor packages
BioC_required_packages <- c("Maaslin2", "ANCOMBC", "biomaRt", "mia", "phyloseq", "microbiome","SIAMCAT")
BioC_required_packages <- BioC_required_packages[!(BioC_required_packages %in% installed.packages())]

if(length(BioC_required_packages) > 0) {
  BiocManager::install(BioC_required_packages[!(BioC_required_packages %in% installed.packages())],
                       ask = FALSE,
                       update = TRUE)
}


# load basic data manipulation packages
library(phyloseq)
library(DT)
library(kableExtra)
library(MultiAssayExperiment)
library(tidyverse)

rm(BioC_required_packages,CRAN_required_packages)
gc(verbose = FALSE)

for(f in list.files("~/bin/biobakeryUtils/", full.names = TRUE)){
  source(f)
}
```

# Directory and data preparation

First thing is clone the repository I made for the Payami Lab and add make sure 
that the `.gitignore` file has exceptions for unwanted data sharing into the 
repository. `NB: This has to be done manually`

```{bash, eval=FALSE}
cd ~

git clone git@github.com:ASAP-MAC/PayamiH.git

cd PayamiH
```

# Work directories and genetic data

1 - Download microbiome data from Zenodo 

**Maker sure you start from the base Git directory**

```{bash setup data directory, eval=FALSE}
# TOPMed data
mkdir -p Data/TOPMed_imputed_genotypes
cp -r /data/project/payami-lab/payamilab/TOPMed_imputed_genotypes_2022/* Data/TOPMed_imputed_genotypes

wget https://zenodo.org/records/7246185/files/Source_Data_24Oct2022.xlsx -O wallen2022_input.xlsx

```

2 - bcftools setup

There are several options. The one proposed below is one of the most universal,
meaning many can set it up.

```{bash setup micromamba, eval=FALSE}
cd ~

curl -Ls https://micro.mamba.pm/api/micromamba/linux-64/latest | tar -xvj micromamba/bin
./bin/micromamba shell init -s bash -r ~/micromamba

micromamba config append channels bioconda

micromamba create -n geneticTools -c bioconda

micromamba activate geneticTools

micromamba install bcftools

```

3 - Indexing of VCF file for fast querying (necessary for bcftools)

```{bash tabix vcf files, eval=FALSE}
cd ~/PayamiH/Data/TOPMed_imputed_genotypes/

for chr in `ls *.dose.vcf.gz`; do sbatch -c1 --mem 8G -pexpress --wrap "bcftools index -c ${chr}"; done
```

4 - Extract samples list to query with `bcftools query`, this is done considering the following:

  1. The only samples needed for the analysis are those beginning with `D`
  2. Keep only samples from point 1 that present across all VCF files. This is 
  redundant but done anyway.

The last step is important for bcftools to work properly in querying samples. 

# Project setup, PD genetics and gut microbiome

```{bash, eval=FALSE}
cd ~/PayamiH

mkdir -p project_1_MicroGenPD/output/

```

# Download all unique GWAS catalog variants with PD (based on `rsID`)

Now we should download the genetic variants that we are interested in. This is the
following method used to retrieve these variants.

  1. search associations using the disease ID `MONDO_0005180`
  
  2. join all useful data together (from step 1)
  
  3. filter for variants with $Pvalue < 5 \times 10^{-8}$ (more than half of these
  do not reach that threshold)
  
  4. Harmonize all variants with `chromosome`, `position`, `risk allele`, ...
  
```{r, eval=!interactive()}
# retrieve associations with GWAS catalog traits
parkinson_gwasCatalog.gwasrapidd <- gwasrapidd::get_associations(efo_id = "MONDO_0005180")

tmp <- list(
  select(parkinson_gwasCatalog.gwasrapidd@associations, 
         association_id, or_per_copy_number, beta_unit, pvalue),
  select(parkinson_gwasCatalog.gwasrapidd@risk_alleles, 
         association_id, variant_id, risk_allele),
  select(parkinson_gwasCatalog.gwasrapidd@genes, 
         association_id, gene_name)) %>% 
  reduce(full_join, by = "association_id") %>% 
  # remove duplicated lines, preferring those where a rik allele has been reported
  arrange(risk_allele) %>%       
  distinct(variant_id, .keep_all = TRUE) %>% 
  # keep only variants that passed nominal Genome-wide significance
  filter(pvalue < 5e-8)

tryCatch(
  load("GWAS_catalog_PD_variants/ensembl.RData"),
  error = function(){
    ensembl <- biomaRt::useEnsembl(biomart = "snps", dataset = "hsapiens_snp")
    save(ensembl, file = "GWAS_catalog_PD_variants/ensembl.RData")
  }
    )


rsid_to_CHR_POS <- biomaRt::getBM(
  attributes = c("refsnp_id", "chr_name", "chrom_start", "chrom_end"), 
      filters = "snp_filter", 
      values = tmp$variant_id, 
      mart = ensembl) %>% 
  rename(
    variant_id = refsnp_id
  ) %>%
  #keep only well defined chromosome scaffolds, including sex chromosome
  filter(
    !startsWith(chr_name, "H")
  ) %>% 
  mutate(
    chr_name = factor(paste0("chr", chr_name), levels = as.character(paste0("chr", c(1:22, "X", "Y"))))
  ) %>% 
  select(
    CHROM = chr_name, 
    CHROM_START = chrom_start, 
    CHROM_END = chrom_end, 
    variant_id
    ) %>% 
  # same trick, to be sure we have only variants with gene names associated, wherever possible
  arrange(rowSums(is.na(.))) %>%       
  distinct(variant_id, .keep_all = TRUE)

parkinson_variants_gwas_catalog.df <- full_join(tmp, rsid_to_CHR_POS, by = "variant_id") %>% 
  select(
    CHROM, CHROM_START, CHROM_END, RSID = variant_id, GENE = gene_name, PD_RISK_ALLELE = risk_allele
  ) %>%
  arrange(CHROM, CHROM_START)

parkinson_variants_gwas_catalog.df$CHROM[parkinson_variants_gwas_catalog.df$RSID == "6:32588205"] <- "chr6"
parkinson_variants_gwas_catalog.df$CHROM_START[parkinson_variants_gwas_catalog.df$RSID == "6:32588205"] <- "32588205"
parkinson_variants_gwas_catalog.df$CHROM_END[parkinson_variants_gwas_catalog.df$RSID == "6:32588205"] <- "32588205"
parkinson_variants_gwas_catalog.df$RSID[parkinson_variants_gwas_catalog.df$RSID == "6:32588205"] <- "rsUnassigned"

datatable(parkinson_variants_gwas_catalog.df, caption = "Unique variants available in the GWAS catalog with P-value , 5E-8. Variants without chromosome and position available will be discarded. Variant with RSID 'rsUnassigned' comes from the original value '6:32588205', which was manually curated to be included into the analysis. Gene annotations were taken 'as is' from the GWAS catalog, to limit variability. Notice that the biomaRt annotation was returning different annotations for < 10% variants.")

```

Unfortunately, despite imputation and manual curation, variant `6:32588205` was
not found in our imputed genotype datasets.

```{r, eval=!interactive()}
library(magrittr)
parkinson_variants_gwas_catalog.df %<>% filter(!is.na(CHROM), !is.na(CHROM_START))

write_tsv(parkinson_variants_gwas_catalog.df, file = "GWAS_catalog_PD_variants/PD_variants_GWAS_catalog.tsv")

for(chr in unique(parkinson_variants_gwas_catalog.df$CHROM)){
  writeLines(text = paste0(filter(parkinson_variants_gwas_catalog.df, CHROM == chr)$CHROM, ":", filter(parkinson_variants_gwas_catalog.df, CHROM == chr)$CHROM_START,"-", filter(parkinson_variants_gwas_catalog.df, CHROM == chr)$CHROM_END), con = paste0("GWAS_catalog_PD_variants/", chr, "_variants_Tabix.txt"))
}

```

```{bash, eval=FALSE}
# from the PayamiH as base directory, do the following
cd Data/GWAS_catalog_PD_variants 

for chr in $(ls | grep -oE "^(chr[0-9]*|chrX)"); do bcftools query -H -f "%CHROM\t%POS\t%REF\t%ALT\t%INFO[\t%GT]\n" -r $(tr "\n" "," < ${chr}_variants_Tabix.txt | sed "s/,$//") ../TOPMed_imputed_genotypes/${chr}.dose.vcf.gz | sed -e "s/#//g" -e "s/:GT//g" -e "s/\[[0-9]*\]//g" -e "s/(null)/INFO/g" > queried_variants_PD/${chr}_pd_Genotypes.txt; done

for chr in $(ls | grep -oE "^(chr[0-9]*|chrX)"); do bcftools query -H -f "%CHROM\t%POS\t%REF\t%ALT\t%INFO[\t%DS]\n" -r $(tr "\n" "," < ${chr}_variants_Tabix.txt | sed "s/,$//") ../TOPMed_imputed_genotypes/${chr}.dose.vcf.gz | sed -e "s/#//g" -e "s/:DS//g" -e "s/\[[0-9]*\]//g" -e "s/(null)/INFO/g" > queried_variants_PD/${chr}_pd_Dosages.txt; done

```

# Data wrapping into `MultiAssayExperiment`

## Prepare microbiome-related data

```{r}

# 1 - metaphlan relative abundances and estimated counts
WallenZD_2022_metaphlan_relab <- readxl::read_xlsx("wallen2022_input.xlsx", sheet = "metaphlan_rel_ab")
WallenZD_2022_metaphlan_counts <- readxl::read_xlsx("wallen2022_input.xlsx", sheet = "metaphlan_counts")
# 1.1 - clean metaphlan tables
WallenZD_2022_metaphlan_relab_wrangled <- wrangle_metaphlan(WallenZD_2022_metaphlan_relab, "Species")
WallenZD_2022_metaphlan_counts_wrangled <- wrangle_metaphlan(WallenZD_2022_metaphlan_counts, "Species")
WallenZD_2022_metaphlan_counts_wrangled$profiles <- round(WallenZD_2022_metaphlan_counts_wrangled$profiles, 0)

# 2 - HUMaNN profiles
# 2.1 KO tables
WallenZD_2022_humann_KO <- readxl::read_xlsx("wallen2022_input.xlsx", sheet = "humann_KO_group_counts") %>% 
  wrangle_humann(input_type = "ko")
WallenZD_2022_humann_KO$profiles <- round(WallenZD_2022_humann_KO$profiles, 0)
WallenZD_2022_humann_pathways <- readxl::read_xlsx("wallen2022_input.xlsx", sheet = "humann_pathway_counts") %>% wrangle_humann(input_type = "pathways")

# 2.2 - Pathways
WallenZD_2022_humann_pathways$profiles <- round(WallenZD_2022_humann_pathways$profiles, 0)

# 3 - sample-participant metadata
WallenZD_2022_metadata <- readxl::read_xlsx("wallen2022_input.xlsx", sheet = "subject_metadata") %>% 
  as.data.frame()
rownames(WallenZD_2022_metadata) <- WallenZD_2022_metadata$sample_name

# reorder colnames so they all match the metadata
samples_intersect <- purrr::reduce(list(WallenZD_2022_metaphlan_relab_wrangled$profiles,
WallenZD_2022_metaphlan_counts_wrangled$profiles,
WallenZD_2022_humann_KO$profiles,
WallenZD_2022_humann_pathways$profiles) %>% lapply(colnames), intersect) 

WallenZD_2022_metaphlan_relab_wrangled$profiles <- WallenZD_2022_metaphlan_relab_wrangled$profiles[, samples_intersect]
WallenZD_2022_metaphlan_counts_wrangled$profiles <- WallenZD_2022_metaphlan_counts_wrangled$profiles[, samples_intersect]
WallenZD_2022_humann_KO$profiles <- WallenZD_2022_humann_KO$profiles[, samples_intersect]
WallenZD_2022_humann_pathways$profiles <- WallenZD_2022_humann_pathways$profiles[, samples_intersect]

sample_metadata.df <- WallenZD_2022_metadata[samples_intersect,] %>% 
  mutate(seqs_scaled = scale(total_sequences))

```

## Prepare genotype data

Both dosages and hardcalls.

```{r load variants dosages and hardcalls}
gwasCatalog_variants_dosages_All <- data.table::fread("GWAS_catalog_PD_variants/PD_variants_GWAS_catalog.tsv", data.table = FALSE) %>% 
  rename(POS = CHROM_START,
         PD.RISK.ALLELE = PD_RISK_ALLELE) %>% 
  select(-CHROM_END) %>% 
  as.data.frame()

gwasCatalog_variants_dosages_in_TOPMed.df <- list.files(path = "GWAS_catalog_PD_variants/queried_variants_PD/", pattern = "*Dosages.txt", full.names = T) %>% 
  sapply(data.table::fread, USE.NAMES = TRUE, simplify = FALSE) %>%
  lapply(inner_join, gwasCatalog_variants_dosages_All, by = c("CHROM", "POS")) %>% 
  lapply(relocate, PD.RISK.ALLELE, RSID, GENE, .after = ALT) %>% 
  purrr::reduce(rbind) %>% 
  # wrangle INFO column for further info
  separate_rows(INFO, sep = ";") %>%
  separate(INFO, into = c("key", "value"), sep = "=", fill = "right") %>%
  pivot_wider(names_from = key, values_from = value) %>%
  mutate(across(everything(), ~replace_na(.x, NA)),
         IMPUTED = ifelse(
           is.na(ER2), "TRUE", "FALSE"), 
         TYPED = NULL) %>% 
  
  relocate(
    INFO_AF = AF, INFO_MAF = MAF, INFO_R2 = R2, INFO_ER2 = ER2, INFO_IMPUTED = IMPUTED, 
    .after = GENE
  ) %>% 
  # as last step to be able to assign row names 2 chunks later
  as.data.frame()


gwasCatalog_variants_hardcalls_in_TOPMed.df <- list.files(path = "GWAS_catalog_PD_variants/queried_variants_PD/", pattern = "*Genotypes.txt", full.names = T) %>% 
  sapply(data.table::fread, USE.NAMES = TRUE, simplify = FALSE) %>% 
  lapply(inner_join, gwasCatalog_variants_dosages_All, by = c("CHROM", "POS")) %>% 
  lapply(relocate, PD.RISK.ALLELE, RSID, GENE, .after = ALT) %>% 
  purrr::reduce(rbind) %>% 
  # wrangle INFO column for further info
  separate_rows(INFO, sep = ";") %>%
  separate(INFO, into = c("key", "value"), sep = "=", fill = "right") %>%
  pivot_wider(names_from = key, values_from = value) %>%
  mutate(across(everything(), ~replace_na(.x, NA)),
         IMPUTED = ifelse(
           is.na(ER2), "TRUE", "FALSE"), 
         TYPED = NULL) %>% 
  
  relocate(
    INFO_AF = AF, INFO_MAF = MAF, INFO_R2 = R2, INFO_ER2 = ER2, INFO_IMPUTED = IMPUTED, 
    .after = GENE
  ) %>% 
  # as last step to be able to assign row names 2 chunks later
  as.data.frame()

# summary table of variants found in the dataset
nVars_by_genes_All <- gwasCatalog_variants_dosages_All %>% 
  mutate(
    `SNCA or HLA` = grepl("SNCA|HLA", GENE), 
    `LRRK2 or GBA` = grepl("LRRK2|GBA", GENE), 
    OTHER = !grepl("SNCA|HLA|GBA|LRRK2", GENE)) %>% 
  .[,(ncol(.)-2):ncol(.)] %>% 
  colSums()

nVars_by_genes_TOPMed <- gwasCatalog_variants_dosages_in_TOPMed.df %>% 
  mutate(
    `SNCA or HLA` = grepl("SNCA|HLA", GENE), 
    `LRRK2 or GBA` = grepl("LRRK2|GBA", GENE), 
    OTHER = !grepl("SNCA|HLA|GBA|LRRK2", GENE)) %>% 
  .[,(ncol(.)-2):ncol(.)] %>% 
  colSums()

cbind(c("Tot" = sum(nVars_by_genes_All), nVars_by_genes_All), 
      c("Tot" = sum(nVars_by_genes_TOPMed), nVars_by_genes_TOPMed)) %>% 
  
  magrittr::set_colnames(c("GWAS Catalog", "TOPMed imputed PayamiLab samples")) %>% 
  kableExtra::kbl(caption = "Comparison of genetic variants associated with PD found in the GWAS catalog and in TOPMed imputed genotypes") %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::add_header_above(c(" " = 1, "N. unique variants associated with PD" = 2))

```

The sample names in the TOPMed data are duplicated with the 
`sample1234_sample1234`, while microbiome data sets have only `sample1234`.

```{r}

variant_metadata_cols <- colnames(gwasCatalog_variants_dosages_in_TOPMed.df)[1:max(which(grepl("INFO", colnames(gwasCatalog_variants_dosages_in_TOPMed.df))))]
sampleIDs <- colnames(gwasCatalog_variants_dosages_in_TOPMed.df)[grepl(paste(paste0("^", rownames(sample_metadata.df), collapse = "|")), colnames(gwasCatalog_variants_dosages_in_TOPMed.df))]

newColnames <- c(variant_metadata_cols, sampleIDs)
names(newColnames) <- c(variant_metadata_cols, sampleIDs %>% strsplit("_") %>% sapply("[", 1))

gwasCatalog_variants_hardcalls_in_TOPMed.df <- select(gwasCatalog_variants_hardcalls_in_TOPMed.df, all_of(newColnames))
gwasCatalog_variants_dosages_in_TOPMed.df <- select(gwasCatalog_variants_dosages_in_TOPMed.df, all_of(newColnames))

```

At this step, we have generated a set of genotypes to test against the microbiome
as found in the GWAS catalog and our genotypes. It is a `samples x genotypes` matrix.

## Generate a `MultiAssayExperiment` object

```{r}
# manipulate genetics data a bit, to make it more usable in the SummarizedExperiment
# context

tmp <- gwasCatalog_variants_dosages_in_TOPMed.df
samples_intersect <- intersect(colnames(tmp), rownames(sample_metadata.df))
variants_made_unique <- make.unique(tmp$RSID)
rownames(tmp) <- variants_made_unique

genetics_matrix_dosages <- as.matrix(tmp[variants_made_unique, samples_intersect])

tmp <- gwasCatalog_variants_hardcalls_in_TOPMed.df
samples_intersect <- intersect(colnames(tmp), rownames(sample_metadata.df))
variants_made_unique <- make.unique(tmp$RSID)
rownames(tmp) <- variants_made_unique

genetics_matrix_hardcalls <- as.matrix(tmp[variants_made_unique, samples_intersect])


genetics_metadata <- tmp[variants_made_unique, variant_metadata_cols]
sample_metadata_genetics <- sample_metadata.df[samples_intersect,]

library(MultiAssayExperiment, quietly = TRUE, verbose = FALSE)

summarizedExperiments.list <- list(
  genotypes_dosages.se = SummarizedExperiment(
    assays = list("genotypes_dosages" = genetics_matrix_dosages, "genotypes_hardcalls" = genetics_matrix_hardcalls),
    rowData = genetics_metadata,
  colData = sample_metadata_genetics
  ),
  
  genotypes_hardcalls.se = SummarizedExperiment(
    assays = list("genotypes_hardcalls" = genetics_matrix_hardcalls),
    rowData = genetics_metadata,
  colData = sample_metadata_genetics
  ), 
  
  metaphlan_RelAbund.se = SummarizedExperiment(
    assays = list("metaphlan_RelAbund" = WallenZD_2022_metaphlan_relab_wrangled$profiles),
    rowData =  WallenZD_2022_metaphlan_relab_wrangled$taxonomies,
    colData = sample_metadata.df
  ),
  
  metaphlan_Counts.se = SummarizedExperiment(
    assays = list("metaphlan_Counts" = WallenZD_2022_metaphlan_counts_wrangled$profiles),
    rowData =  WallenZD_2022_metaphlan_counts_wrangled$taxonomies,
    colData = sample_metadata.df
  ),
  
  humann_KO.se = SummarizedExperiment(
    assays = list("humann_KO" = WallenZD_2022_humann_KO$profiles),
    rowData =  WallenZD_2022_humann_KO$description,
    colData = sample_metadata.df
  ),
  
  humann_pathways.se = SummarizedExperiment(
    assays = list("humann_pathways" = WallenZD_2022_humann_pathways$profiles),
    rowData =  WallenZD_2022_humann_pathways$description,
    colData = sample_metadata.df
  )
)

assays_vec <- lapply(summarizedExperiments.list, colnames) %>% 
  lapply(length)

assays_vec <- lapply(names(assays_vec), function(x) rep(x, assays_vec[[x]])) %>% 
  purrr::reduce(c) %>% 
  factor(levels = names(assays_vec))

assays_sample_map <- data.frame(
  assay = assays_vec,
  primary = lapply(summarizedExperiments.list, colnames) %>% purrr::reduce(c),
  colname = lapply(summarizedExperiments.list, colnames) %>% purrr::reduce(c)
)

WallenZD_2022_data.mae <- MultiAssayExperiment(
  experiments = summarizedExperiments.list,
  sampleMap = assays_sample_map,
  colData = sample_metadata.df
)
```

```{r clean environment and memory for next analyses}

save(WallenZD_2022_data.mae, file = "WallenZD_2022_data.mae.RData")

```

# Final data object generated - a quick look

```{r}
WallenZD_2022_data.mae
```

# Session Info

```{r}
sessionInfo()
```

